---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# marlin

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://www.tidyverse.org/lifecycle/#experimental)
[![CRAN status](https://www.r-pkg.org/badges/version/mar)](https://CRAN.R-project.org/package=mar)
<!-- badges: end -->

marlin is a package for efficiently running simulations of marine fauna and fisheries. Age-structured population model of different (independent) animal types in a 2D system with multiple fishing fleets. 

## Installation

You can install the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("DanOvando/marlin")
```
## Naviation

The core wrapper function is located in R/simmar.R. This funcion keeps track of each of the populations and fleets. 

The actual population models are found in src/fish_model.cpp. Additional modules will be put in there as they are developed

### Troubleshooting

Make sure you try the install with a fresh R session (go to "Session>Restart R" to make sure)

If you run into an error, first off try updating your R packages. From there....

If your version of R is lower than 3.5, you might want to consider updating R itself. Updating from 3.51 to 3.52 shouldn't be any hassle. BIG WARNING THOUGH, updating from say R 3.1 to 3.5 is a major update, and you'll lose all your installed packages in the process. I recommend following the instructions [here](https://www.datascienceriot.com/r/upgrade-R-packages/) to deal with that, but even with that fix it can take a while, so I don't recommend doing a major R update if you're on a deadline. There are also packages to help you with this process, specifically [`installR`](https://github.com/talgalili/installr/issues) for Windows and [`updateR`](https://github.com/AndreaCirilloAC/updateR) for Mac. 

From there...

- On Windows, make sure you have the appropriate version of Rtools installed ([here](https://cran.r-project.org/bin/windows/Rtools/)), most likely Rtools35 if you have R version 3.3 or higher
  - Make sure that you select the box that says something about adding Rtools to the PATH variable

- On macOS, there might be some issues with the your compiler, particularly if your version of R is less than 4.0.0.

If you get an error that says something like `clang: error: unsupported option '-fopenmp'`, follow the instructions [here](https://thecoatlessprofessor.com/programming/cpp/r-compiler-tools-for-rcpp-on-macos-before-r-4.0.0/)

Once you've tried those, restart your computer and try running

Below are a bunch of examples showing what `marlin` can do

## Simple Example

Let's start with a simple one-fleet one-critter example to illustrate the various options in `marlin`


```{r}
library(marlin)
library(tidyverse)
options(dplyr.summarise.inform = FALSE)
theme_set(marlin::theme_marlin())

resolution <- 25 # resolution is in squared patches, so 20 implies a 20X20 system, i.e. 400 patches 

years <- 20

seasons <- 1

time_step <- 1 / seasons

steps <- years * seasons

fauna <- 
  list(
    "bigeye" = create_critter(
      common_name = "bigeye tuna",
      adult_movement = 1,
      adult_movement_sigma = 10,
      rec_form = 1,
      seasons = seasons,
      fished_depletion = .5
    )
  )

# create a fleets object, which is a list of lists (of lists). Each fleet has one element, 
# with lists for each species inside there. Price specifies the price per unit weight of that 
# species for that fleet
# sel_form can be one of logistic or dome


fleets <- list(
  "longline" = create_fleet(
    list("bigeye" = Metier$new(
        critter = fauna$bigeye,
        price = 10,
        sel_form = "logistic",
        sel_start = 1,
        sel_delta = .01,
        catchability = 0,
        p_explt = 1
      )
    ),
    base_effort = resolution ^ 2
  )
)

a <- Sys.time()

fleets <- tune_fleets(fauna, fleets, tune_type = "depletion") 

Sys.time() - a


fauna$bigeye$plot()

a <- Sys.time()

sim <- simmar(fauna = fauna,
                  fleets = fleets,
                  years = years)

Sys.time() - a
  

```


we can then use `process_marlin` and `plot_marlin` to examine the simulation


```{r}

processed_marlin <- process_marlin(sim = sim, time_step = time_step)

plot_marlin(processed_marlin)

plot_marlin(processed_marlin, plot_var = "c")

plot_marlin(processed_marlin, plot_var = "n", plot_type = "length", fauna = fauna)

plot_marlin(processed_marlin, plot_var = "ssb", plot_type = "space")


```


## Simple Example: seasonal habitat and movement and spatial q

Now let's add in different adult habitats by season, different movement rates by season, specified recruitment habitat, and a spatial dimension to catchability, and quarterly time steps

```{r}

seasons <- 4

time_step <-  1 / seasons

bigeye_habitat <- expand_grid(x = 1:resolution, y = 1:resolution) %>%
  mutate(habitat =  dnorm((x ^ 2 + y ^ 2), 300, 100)) %>% 
  pivot_wider(names_from = y, values_from = habitat) %>% 
  select(-x) %>% 
  as.matrix()


bigeye_habitat2 <- expand_grid(x = 1:resolution, y = 1:resolution) %>%
  mutate(habitat =  dnorm((x ^ .2 + y ^ .2), 100, 100)) %>% 
  pivot_wider(names_from = y, values_from = habitat) %>% 
  select(-x) %>% 
  as.matrix()

bigeye_q <- expand_grid(x = 1:resolution, y = 1:resolution) %>%
  dplyr::mutate(habitat = rlnorm(resolution^2)) %>% 
  pivot_wider(names_from = y, values_from = habitat) %>% 
  select(-x) %>% 
  as.matrix()


fauna <- 
  list(
    "bigeye" = create_critter(
      common_name = "bigeye tuna",
      seasonal_habitat = list(bigeye_habitat,bigeye_habitat2),
      season_blocks = list(c(1,2),c(3,4)),
     adult_movement = list(c(0,0),c(10,10)),# the mean number of patches moved by adults
      adult_movement_sigma = list(c(2,2), c(.1,.1)), # standard deviation of the number of patches moved by adults
      rec_form = 2,
      seasons = seasons,
      init_explt =  .1,
     explt_type = "f"
    )
  )

# create a fleets object, which is a list of lists (of lists). Each fleet has one element, 
# with lists for each species inside there. Price specifies the price per unit weight of that 
# species for that fleet
# sel_form can be one of logistic or dome


fleets <- list(
  "longline" = create_fleet(
    list("bigeye" = Metier$new(
        critter = fauna$bigeye,
        price = 10,
        sel_form = "logistic",
        sel_start = 1,
        sel_delta = .01,
        catchability = 1e-3,
        p_explt = 1,
        spatial_catchability = bigeye_q
      )
    ),
    base_effort = resolution ^ 2
  )
)

a <- Sys.time()

fleets <- tune_fleets(fauna, fleets) 

Sys.time() - a


fauna$bigeye$plot()

a <- Sys.time()

sim2 <- simmar(fauna = fauna,
                  fleets = fleets,
                  years = years)

Sys.time() - a
  

processed_marlin <- process_marlin(sim = sim2, time_step = time_step)

plot_marlin(processed_marlin)

plot_marlin(processed_marlin, plot_var = "c")

plot_marlin(processed_marlin, plot_var = "n", plot_type = "length", fauna = fauna)

plot_marlin(processed_marlin, plot_var = "ssb", plot_type = "space")



```


## Two Species and two fleets with bells and whistles


```{r example}

seasons <- 4

steps <- years * seasons

time_step <- 1 / seasons
# for now make up some habitat


skipjack_habitat <- expand_grid(x = 1:resolution, y = 1:resolution) %>%
  dplyr::mutate(habitat =  dnorm((x ^ 2 + y ^ 2), 20, 200)) %>% 
  pivot_wider(names_from = y, values_from = habitat) %>% 
  select(-x) %>% 
  as.matrix()


bigeye_habitat <- expand_grid(x = 1:resolution, y = 1:resolution) %>%
  mutate(habitat =  dnorm((x ^ 2 + y ^ 2), 300, 100)) %>% 
  pivot_wider(names_from = y, values_from = habitat) %>% 
  select(-x) %>% 
  as.matrix()


bigeye_habitat2 <- expand_grid(x = 1:resolution, y = 1:resolution) %>%
  mutate(habitat =  dnorm((x ^ .2 + y ^ .2), 100, 100)) %>% 
  pivot_wider(names_from = y, values_from = habitat) %>% 
  select(-x) %>% 
  as.matrix()

a <- Sys.time()

fauna <- 
  list(
    "skipjack" = create_critter(
      scientific_name = "Katsuwonus pelamis",
      seasonal_habitat = list(skipjack_habitat, skipjack_habitat), # pass habitat as lists
      season_blocks = list(c(1, 2), c(3, 4)), # seasons each habitat apply to
      recruit_habitat = skipjack_habitat,
      adult_movement = 2,# the mean number of patches moved by adults
      adult_movement_sigma = 2, # standard deviation of the number of patches moved by adults
      fished_depletion = .6, # desired equilibrium depletion with fishing (1 = unfished, 0 = extinct),
      rec_form = 1, # recruitment form, where 1 implies local recruitment
      seasons = seasons,
      init_explt = 0.2, 
      explt_type = "f"
      ),
    "bigeye" = create_critter(
      common_name = "bigeye tuna",
      seasonal_habitat = list(bigeye_habitat, bigeye_habitat2), # pass habitat as lists
      season_blocks = list(c(1, 2), c(3, 4)), # seasons each habitat apply to
      recruit_habitat = bigeye_habitat,
      adult_movement = 3,
      adult_movement_sigma = 1,
      fished_depletion = .1,
      rec_form = 1,
      seasons = seasons,
      init_explt = 0.3, 
      explt_type = "f"
    )
  )
Sys.time() - a

# create a fleets object, which is a list of lists (of lists). Each fleet has one element, 
# with lists for each species inside there. Price specifies the price per unit weight of that 
# species for that fleet
# sel_form can be one of logistic or dome


fleets <- list(
  "longline" = create_fleet(
    list(
       "skipjack" = Metier$new(
        critter = fauna$skipjack,
        price = 100,
        # price per unit weight
        sel_form = "logistic",
        # selectivity form, one of logistic or dome
        sel_start = .3,
        # percentage of length at maturity that selectivity starts
        sel_delta = .1,
        # additional percentage of sel_start where selectivity asymptotes
        catchability = .01,
        # overwritten by tune_fleet but can be set manually here
        p_explt = 1
      ),
       "bigeye" = Metier$new(
        critter = fauna$bigeye,
        price = 10,
        sel_form = "logistic",
        sel_start = .1,
        sel_delta = .01,
        catchability = 0,
        p_explt = 1
      )
    ),
    base_effort = resolution ^ 2
  ),
  "purseseine" = create_fleet(list(
    skipjack = Metier$new(
      critter = fauna$skipjack,
      price = 100,
      sel_form = "logistic",
      sel_start = 0.25,
      sel_delta = .1,
      catchability = .01,
      p_explt = 0.9
    ),
    bigeye = Metier$new(
      critter = fauna$bigeye,
      price = 100,
      sel_form = "logistic",
      sel_start = .25,
      sel_delta = .5,
      catchability = .01,
      p_explt = 1
    )
  ),
  base_effort = resolution ^ 2)
)

a <- Sys.time()

fleets <- tune_fleets(fauna, fleets) 

Sys.time() - a


# run simulations

# run the simulation using marlin::simmar
a <- Sys.time()

sim3 <- simmar(fauna = fauna,
                  fleets = fleets,
                  years = years)

Sys.time() - a
  
processed_marlin <- process_marlin(sim = sim3, time_step = time_step)

plot_marlin(processed_marlin)

plot_marlin(processed_marlin, plot_var = "c")

plot_marlin(processed_marlin, plot_var = "n", plot_type = "length", fauna = fauna)

plot_marlin(processed_marlin, plot_var = "ssb", plot_type = "space")



```


## Evaluating MPAs

Now let's compare the effect of an MPA on two species


sharks live nearshore


```{r}
library(marlin)
library(tidyverse)

theme_set(marlin::theme_marlin())

resolution <- 20 # resolution is in squared patches, so 20 implies a 20X20 system, i.e. 400 patches 

seasons <- 1

years <- 50

tune_type <- "depletion"

steps <- years * seasons

# for now make up some habitat

yft_habitat <- expand_grid(x = 1:resolution, y = 1:resolution) %>%
  mutate(habitat =  .05 * x) %>% 
  pivot_wider(names_from = y, values_from = habitat) %>% 
  select(-x) %>% 
  as.matrix()
 

mako_habitat <- expand_grid(x = 1:resolution, y = 1:resolution) %>%
  mutate(habitat =  dnorm(x,17,5)) %>% 
  pivot_wider(names_from = y, values_from = habitat) %>% 
  select(-x) %>% 
  as.matrix()


# create a fauna object, which is a list of lists

fauna <- 
  list(
    "Yellowfin Tuna" = create_critter(
      scientific_name = "Thunnus albacares",
      seasonal_habitat = list(yft_habitat), # pass habitat as lists
      recruit_habitat = yft_habitat,
      adult_movement = 0,# the mean number of patches moved by adults
      adult_movement_sigma = 4, # standard deviation of the number of patches moved by adults
      fished_depletion = .4, # desired equilibrium depletion with fishing (1 = unfished, 0 = extinct),
      rec_form = 1, # recruitment form, where 1 implies local recruitment
      seasons = seasons,
      init_explt = 0.12, 
      explt_type = "f"
      ),
    "Shortfin Mako" = create_critter(
      scientific_name = "Isurus oxyrinchus",
      seasonal_habitat = list(mako_habitat), # pass habitat as lists
      recruit_habitat = mako_habitat,
      adult_movement = 5,
      adult_movement_sigma = 3,
      fished_depletion = .3,
      rec_form = 1,
      burn_years = 200,
      seasons = seasons,
      init_explt = .12, 
      explt_type = "f",
      fec_form = "linear",
      weight_a = 2 # average two offspring per shark
    )
  )

fauna$`Yellowfin Tuna`$plot()


# create a fleets object, which is a list of lists (of lists). Each fleet has one element, 
# with lists for each species inside there. Price specifies the price per unit weight of that 
# species for that fleet
# sel_form can be one of logistic or dome

fleets <- list("longline" = create_fleet(list(
  `Yellowfin Tuna` = Metier$new(
    critter = fauna$`Yellowfin Tuna`,
    price = 100, # price per unit weight
    sel_form = "logistic", # selectivity form, one of logistic or dome
    sel_start = .3, # percentage of length at maturity that selectivity starts
    sel_delta = .1, # additional percentage of sel_start where selectivity asymptotes
    catchability = .01, # overwritten by tune_fleet but can be set manually here
    p_explt = 1
    ),
  `Shortfin Mako` = Metier$new(
    critter = fauna$`Shortfin Mako`,
    price = 10,
    sel_form = "logistic",
    sel_start = .1,
    sel_delta = .01,
    catchability = 0,
    p_explt = 1
  )),
  mpa_response = "stay",
  base_effort = resolution^2
))

a <- Sys.time()

fleets <- tune_fleets(fauna, fleets, tune_type = tune_type) # tunes the catchability by fleet to achieve target depletion

Sys.time() - a

# run simulations

a <- Sys.time()

nearshore <- simmar(fauna = fauna,
                  fleets = fleets,
                  years = years)

Sys.time() - a
  
proc_nearshore <- process_marlin(nearshore, time_step =  fauna[[1]]$time_step)

```

create an MPA
```{r}
set.seed(42)
#specify some MPA locations
mpa_locations <- expand_grid(x = 1:resolution, y = 1:resolution) %>%
mutate(mpa = x > 15 & y < 15)

mpa_locations %>% 
  ggplot(aes(x,y, fill = mpa)) + 
  geom_tile() + 
  scale_fill_brewer(palette = "Accent", direction  = -1, name = "MPA") + 
  scale_x_continuous(name = "Lat") + 
  scale_y_continuous(name = "Lon")
```

And now apply MPA

```{r}
a <- Sys.time()

nearshore_mpa <- simmar(
  fauna = fauna,
  fleets = fleets,
  years = years,
  mpas = list(locations = mpa_locations,
              mpa_year = floor(years * .5))
)

Sys.time() - a

proc_nearshore_mpa <- process_marlin(nearshore_mpa, time_step =  fauna[[1]]$time_step)

```


Now though, consider a different scenario. Here the tunas still slightly prefer their same nearshore habitat, but now the shortfin mako population primarily lives farther offshore. We will first simulate that population without the MPA, and then assess the effects of the exact same MPA on this new scenario. 


```{r}

yft_habitat <- expand_grid(x = 1:resolution, y = 1:resolution) %>%
  mutate(habitat =  .2 * x) %>% 
  pivot_wider(names_from = y, values_from = habitat) %>% 
  select(-x) %>% 
  as.matrix()
 

mako_habitat <- expand_grid(x = 1:resolution, y = 1:resolution) %>%
  mutate(habitat =  dnorm(x, 9,5)) %>% 
  pivot_wider(names_from = y, values_from = habitat) %>% 
  select(-x) %>% 
  as.matrix()


# create a fauna object, which is a list of lists

fauna <- 
  list(
    "Yellowfin Tuna" = create_critter(
      scientific_name = "Thunnus albacares",
      seasonal_habitat = list(yft_habitat), # pass habitat as lists
      recruit_habitat = yft_habitat,
      adult_movement = 0,# the mean number of patches moved by adults
      adult_movement_sigma = 20, # standard deviation of the number of patches moved by adults
      fished_depletion = .4, # desired equilibrium depletion with fishing (1 = unfished, 0 = extinct),
      rec_form = 1, # recruitment form, where 1 implies local recruitment
      seasons = seasons,
      init_explt = .12, 
      explt_type = "f"
      ),
    "Shortfin Mako" = create_critter(
      scientific_name = "Isurus oxyrinchus",
      seasonal_habitat = list(mako_habitat), # pass habitat as lists
      recruit_habitat = mako_habitat,
      adult_movement = 3,
      adult_movement_sigma = 1,
      fished_depletion = .3,
      rec_form = 1,
      burn_years = 200,
            seasons = seasons,
            init_explt = .12, 
      explt_type = "f",
          fec_form = "linear",
      weight_a = 2
    )
  )

fleets <- tune_fleets(fauna, fleets, tune_type = tune_type) # tunes the catchability by fleet to achieve target depletion

# run simulations

# run the simulation using marlin::simmar
a <- Sys.time()

offshore <- simmar(fauna = fauna,
                  fleets = fleets,
                  years = years)

Sys.time() - a
  
proc_offshore <- process_marlin(offshore, time_step =  fauna[[1]]$time_step)

a <- Sys.time()

offshore_mpa_sim <- simmar(
  fauna = fauna,
  fleets = fleets,
  years = years,
  mpas = list(locations = mpa_locations,
              mpa_year = floor(years * .5))
)

Sys.time() - a


proc_offshore_mpa <- process_marlin(offshore_mpa_sim, time_step =  fauna[[1]]$time_step)

```



```{r}
plot_marlin(
  `MPA: Sharks Offshore` = proc_offshore_mpa,
  `No MPA` = proc_nearshore,
  `MPA: Sharks Nearshore` = proc_nearshore_mpa,
  steps_to_plot = NA,
  plot_var = "ssb"
)

```

```{r}
plot_marlin(
  `MPA: Sharks Offshore` = proc_offshore_mpa,
  `No MPA` = proc_nearshore,
  `MPA: Sharks Nearshore` = proc_nearshore_mpa,
  plot_var = "b",
  plot_type = "space",
  steps_to_plot = c(1,25,50))
```


## Defacto MPAs through bycatch penalties

Now consider a case where prices for shortfin mako are negative creating defacto MPAs


```{r}
years <- 100

tune_type <- "explt"

# make up some habitat

yft_habitat <- expand_grid(x = 1:resolution, y = 1:resolution) %>%
  mutate(habitat =  .05 * x) %>% 
  pivot_wider(names_from = y, values_from = habitat) %>% 
  select(-x) %>% 
  as.matrix()


mako_habitat <- expand_grid(x = 1:resolution, y = 1:resolution) %>%
  mutate(habitat =  x > 12 & y >12) %>% 
  pivot_wider(names_from = y, values_from = habitat) %>% 
  select(-x) %>% 
  as.matrix()


# create a fauna object, which is a list of lists

fauna <- 
  list(
    "Yellowfin Tuna" = create_critter(
      scientific_name = "Thunnus albacares",
      seasonal_habitat = list(yft_habitat), 
      recruit_habitat = yft_habitat,
      adult_movement = 0,
      adult_movement_sigma = 4, 
      fished_depletion = .4, 
      rec_form = 1, 
      seasons = seasons,
      init_explt = 0.12, 
      explt_type = "f"
    ),
    "Shortfin Mako" = create_critter(
      scientific_name = "Isurus oxyrinchus",
      seasonal_habitat = list(mako_habitat), 
      recruit_habitat = mako_habitat,
      adult_movement = 5,
      adult_movement_sigma = 3,
      fished_depletion = .3,
      rec_form = 1,
      burn_years = 200,
      seasons = seasons,
      init_explt = 0.1, 
      explt_type = "f"
    )
  )

# create a fleets object, accounting a negative price to shortfin makos

fleets <- list("longline" = create_fleet(list(
  `Yellowfin Tuna` = Metier$new(
    critter = fauna$`Yellowfin Tuna`,
    price = 100, # price per unit weight
    sel_form = "logistic", # selectivity form, one of logistic or dome
    sel_start = .3, # percentage of length at maturity that selectivity starts
    sel_delta = .1, # additional percentage of sel_start where selectivity asymptotes
    catchability = .01, # overwritten by tune_fleet but can be set manually here
    p_explt = 1
  ),
  `Shortfin Mako` = Metier$new(
    critter = fauna$`Shortfin Mako`,
    price = -20000,
    sel_form = "logistic",
    sel_start = .1,
    sel_delta = .01,
    catchability = 0,
    p_explt = 1
  )),
  mpa_response = "stay",
  base_effort = resolution^2
))

a <- Sys.time()

fleets <- tune_fleets(fauna, fleets, tune_type = tune_type) # tunes the catchability by fleet to achieve target depletion

Sys.time() - a

# run simulations

# run the simulation using marlin::simmar
negative_prices <- simmar(fauna = fauna,
                          fleets = fleets,
                          years = years)

proc_negative_prices <- process_marlin(negative_prices, time_step =  fauna[[1]]$time_step)
```


```{r}
plot_marlin(
  `De-Facto MPA` = proc_negative_prices,
  plot_var = "ssb",
  plot_type = "space",
  steps_to_plot = c(1,25,50))
```




