---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  message = FALSE,
  warning = FALSE
)
set.seed(42)
```

# marlin
<!-- badges: start -->
<!-- [![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://www.tidyverse.org/lifecycle/#experimental) -->
<!-- [![CRAN status](https://www.r-pkg.org/badges/version/mar)](https://CRAN.R-project.org/package=mar) -->
<!-- badges: end -->

This package is back under active construction and so many of the vignettes are still works-in-progress at this time. 

`marlin` is a package for efficiently running simulations of marine fauna and fisheries. It can track the age-structured populations of multiple independent animals targeted by multiple fishing fleets across a two dimensional space. It allows users to simulate a wide range of life histories, movement dynamics, fishing fleet behaviors, and management strategies. See examples below and vignettes under the articles tab for how to use `marlin` to do things like simulate... 
  -  the impacts of displaced fishing effort across multiple species
  -  seasonal spawning aggregations or climate induced range shifts
  -  the impacts of port distance in fleet dynamics
  -  optimized marine protected area network design
  -  Data-generating processes for spatial-stock assessment

The model is described in Ovando *et al.* (2023)  

Ovando, Daniel, Darcy Bradley, Echelle Burns, Lennon Thomas, and James Thorson. “Simulating Benefits, Costs and Trade-Offs of Spatial Management in Marine Social-Ecological Systems.” Fish and Fisheries, (2023). https://doi.org/10.1111/faf.12804.


  
## What is `marlin` for?

`marlin` is best suited for asking *what if* rather that *what will* questions. 

In other words, `marlin` is better suited to asking "how might the impacts of my proposed MPA network change if fishing effort is displaced rather than removed?" rather than "*what will* the impacts of my proposed MPA network be on biodiversity down to the 6th decimal place". 

This is because `marlin` is a *structural* rather than *statistical* simulation model, meaning the parameters in the model cannot be directly fit to explain patterns in observed data. Rather, users must manually adjust the parameters of the model to reflect the general dynamics of the system they are interested in. 

This means that "how much of an impact does hyperallometry have on MPA outcomes" is a much simpler question than "what will be the catch and biodiversity impacts of the MPA I am designing for 17 different data-limited species in a rapidly changing small bay". 

Both are doable, and `marlin` can provide general insights in both cases, but the more specific and *what will* style of question you ask, the harder it will be to ensure the model is appropriately set up. 

Some use cases we envision for marlin are

- Assessing sensitivity of MPA network designs to key ecological and economic unknowns

- Design of dynamic ocean management strategies under climate change

- Management strategy evaluation of spatially-explicit fisheries management 

## Installation

You can install the development version from [GitHub](https://github.com/) using:

``` r
# install.packages("devtools")
devtools::install_github("DanOvando/marlin")
```


### Installation Troubleshooting


Make sure you try the install with a fresh R session (go to "Session>Restart R" to make sure)

If you run into an error, first off try updating your R packages. From there....

If your version of R is lower than 4.0, you might want to consider updating R itself. 

In order to install and run `marlin` you need to have R compiler tools correctly set up. 

To do this... 

- On Windows, make sure you have the appropriate version of Rtools installed ([here](https://cran.r-project.org/bin/windows/Rtools/)), most likely Rtools4X if you have R version 4.X
    - Make sure that you select the box that says something about adding Rtools to the PATH variable if prompted

- On macOS I recommend either using the unofficial [macrtools](https://github.com/coatless-mac/macrtools) package, or if you prefer you can follow the official instructions on CRAN [here](https://mac.r-project.org/tools/). If you're using a version of R pre- 4.XX, then follow instructions [here](https://thecoatlessprofessor.com/programming/cpp/r-compiler-tools-for-rcpp-on-macos-before-r-4.0.0/), though your mileage may vary depending on how out of date you R ecosystem is. 

Once you've got your R compiler tools installed correctly, close down then re-open R and then try installing `marlin` again.


## A Caveat

`marlin` is designed to be fast and relatively user friendly, at least for the complexity of the modeling it enables users to do. However, it is not a tool for fully automated MPA design; users will still need to be able to understand the core functionality of the model, and mileage with the model will depend greatly on the creativity and R skills of the user. 

In particular, while we provide some wrappers for organizing common outputs and plots, `marlin` produces far more outputs, and is designed for more use cases, than we can preemptively plan for. 

Given that `marlin` tracks populations and fleet dynamics across ages, time steps, space, species, and fleet units, the outputs are rather unruly. Skills in wrangling list objects will serve you will, along with tidy-data manipulation. [`purrr`](https://purrr.tidyverse.org/) is really really good at this. 

You do not need the `tidyverse` to run `marlin`, but you will see that we make extensive use of it, in particular `dplyr`, `tidyr`, `ggplot2` and `purrr` in the examples below, to extract useful results from `marlin`. 

## Simple Example

Let's start with a simple one-fleet one-critter example to illustrate the various options in `marlin`. 

Some of the core options for `marlin` 

- `resolution`: the number of patches to each a side of the 2D  grid. Setting one number creates and X by X matrix( e.g. resolution = 10 creates a 10x10 simulation grid). Alternatively you can specify a vector of length two showing the number of dimensions in the X and Y dimension. So, `resolution = c(2,10)` creates a 2 x 10 simulation grid. 

- `years`: the number of years to run the simulation

- `seasons`: the number of seasons per year. 2 would mean that each year is divided into two time steps, 4 would mean a quarterly model, 12 a monthly model, etc. 


```{r}
library(marlin)
library(tidyverse)
options(dplyr.summarise.inform = FALSE)
theme_set(marlin::theme_marlin(base_size = 42))

resolution <- c(5, 10) # resolution is in squared patches, so 20 implies a 20X20 system, i.e. 400 patches
years <- 20

seasons <- 4

time_step <- 1 / seasons

steps <- years * seasons
```


From there, `marlin` works be specifying three list objects

- `fauna`: a list object of the life histories of the different species being simulated

- `fleets`: a list object of the fleet dynamics to be simulated

- `manager`: a list of the management strategies being employed

### A Warning about Default Life History Values
The `fauna` object is a list with individual slots for every "critter" you want to simulate, created by the `create_critter` function. If you supply a common or scientific name, the model will try and populate the life history values for your species based on the values reported in [FishLife](https://github.com/James-Thorson-NOAA/FishLife). Note that you **NEED** an internet connection for this feature to work. Running `marlin` offline requires you to supply all life history values on your own. 

**WARNING**. The values from `FishLife` are model estimates. You should always check the results to see if they are good enough for your application, and where possible / needed update the values with more precise local studies. 

When you use `FishLife` to look up your life history values the objecting resulting from `create_critter` will have an entry called `closest_taxa_match`. This is the taxonomic level that `FishLife` ended up using given the scientific or common name supplied to it. You should double check that the result is as close to the species you wanted as possible!

We recommend using `scientific_name` rather than `common_name`. If you tried to look up by common_name and get an error that says `Error in seq.default(min_age, max_age, by = time_step)...` it's most likely a database connection problem and simply running the code again will work. 

### Creating a basic simulation
```{r}
fauna <-
  list(
    "bigeye" = create_critter(
      common_name = "bigeye tuna",
      adult_diffusion = 10,
      density_dependence = "local_habitat",
      seasons = seasons,
      fished_depletion = .25,
      resolution = resolution,
      steepness = 0.6,
      ssb0 = 42,
      m = 0.4
    )
  )

fauna$bigeye$plot()

resolution <- fauna[[1]]$resolution
```


The `fleets` object is a list of individual fishing fleets created by the `create_fleet` function. Importantly, each fleet is broken up into metiers specifying the dynamics of a fishing fleet with regards to a matching critter in the `fauna` object. So, in this case we have a longline fleet, with a metier for bigeye tuna specifying things like the price per unit weight, selectivity, and p_explt. `p_explt` tracks the relative exploitation of each metier in the total exploitation of each species. So, if one metier has a `p_explt` of 1 for species X, and another a `p_explt` of 2 for the same species, 50% of the mortality for species X comes from the first metier (p_explt values are relative, not absolute).

```{r}
fleets <- list(
  "longline" = create_fleet(
    list("bigeye" = Metier$new(
      critter = fauna$bigeye,
      price = 10,
      sel_form = "logistic",
      sel_start = 1,
      sel_delta = .01,
      catchability = 0,
      p_explt = 1,
    )),
    base_effort = prod(resolution),
    resolution = resolution
  )
)
```

We then use the `tune_fleets` function as needed to tune the dynamics of the fleet to achieve specific objectives. In this case, we specified a target equilibrium fished depletion of 25% (meaning biomass at equilibrium is 25% of unfished biomass). `tune_fleets` adjusts the catchability coefficients by fleet to achieve the desired depletion level, taking into account the dynamics and `p_explt` values per metier. 

```{r}
fleets <- tune_fleets(fauna, fleets, tune_type = "depletion")
```


From there, we run the simulation by passing the `fauna` and `fleet` options to the `simmar` function


```{r}
start_time <- Sys.time()

example_sim <- simmar(
  fauna = fauna,
  fleets = fleets,
  years = years
)

Sys.time() - start_time
```


we can then use `process_marlin` and `plot_marlin` to examine the simulation


```{r}
processed_marlin <- process_marlin(sim = example_sim, time_step = time_step)

plot_marlin(processed_marlin)

plot_marlin(processed_marlin, plot_var = "c", max_scale = FALSE)

plot_marlin(processed_marlin, plot_var = "n", plot_type = "length", fauna = fauna)

plot_marlin(processed_marlin, plot_var = "ssb", plot_type = "space", steps_to_plot = max(processed_marlin$fauna$step))
```


## Two Species and two fleets with bells and whistles

Now, let's make this a two species and two fleet example with some seasonal dynamics, where the species have different habitats in different seasons. 

```{r example}
seasons <- 4

steps <- years * seasons

time_step <- 1 / seasons

skipjack_diffusion <- 2

bigeye_diffusion <- 5

# for now make up some habitat

skipjack_habitat <- expand_grid(x = 1:resolution[1], y = 1:resolution[2]) %>%
  dplyr::mutate(
    habitat = dnorm((x^2 + y^2), 20, 200),
    habitat = habitat / max(habitat) * skipjack_diffusion
  ) |>
  pivot_wider(names_from = x, values_from = habitat) %>%
  select(-y) %>%
  as.matrix()


bigeye_habitat <- expand_grid(x = 1:resolution[1], y = 1:resolution[2]) %>%
  mutate(
    habitat = dnorm((x^2 + y^2), 300, 100),
    habitat = habitat / max(habitat) * bigeye_diffusion
  ) %>%
  pivot_wider(names_from = x, values_from = habitat) %>%
  select(-y) %>%
  as.matrix()


bigeye_habitat2 <- expand_grid(x = 1:resolution[1], y = 1:resolution[2]) %>%
  mutate(
    habitat = dnorm((x^.2 + y^.2), 100, 100),
    habitat = habitat / max(habitat) * bigeye_diffusion
  ) %>%
  pivot_wider(names_from = x, values_from = habitat) %>%
  select(-y) %>%
  as.matrix()

a <- Sys.time()

fauna <-
  list(
    "skipjack" = create_critter(
      scientific_name = "Katsuwonus pelamis",
      habitat = list(skipjack_habitat, skipjack_habitat), # pass habitat as lists
      season_blocks = list(c(1, 2), c(3, 4)), # seasons each habitat apply to
      recruit_habitat = skipjack_habitat,
      adult_diffusion = skipjack_diffusion, # standard deviation of the number of patches moved by adults
      fished_depletion = .6, # desired equilibrium depletion with fishing (1 = unfished, 0 = extinct),
      density_dependence = "global_habitat", # recruitment form, where 1 implies local recruitment
      seasons = seasons,
      init_explt = 0.2,
      explt_type = "f"
    ),
    "bigeye" = create_critter(
      common_name = "bigeye tuna",
      habitat = list(bigeye_habitat, bigeye_habitat2), # pass habitat as lists
      season_blocks = list(c(1, 2), c(3, 4)), # seasons each habitat apply to
      recruit_habitat = bigeye_habitat,
      adult_diffusion = bigeye_diffusion,
      fished_depletion = .1,
      density_dependence = "local_habitat",
      seasons = seasons,
      init_explt = 0.3,
      explt_type = "f"
    )
  )
Sys.time() - a

# create a fleets object, which is a list of lists (of lists). Each fleet has one element,
# with lists for each species inside there. Price specifies the price per unit weight of that
# species for that fleet
# sel_form can be one of logistic or dome


fleets <- list(
  "longline" = create_fleet(
    list(
      "skipjack" = Metier$new(
        critter = fauna$skipjack,
        price = 100,
        # price per unit weight
        sel_form = "logistic",
        # selectivity form, one of logistic or dome
        sel_start = .3,
        # percentage of length at maturity that selectivity starts
        sel_delta = .1,
        # additional percentage of sel_start where selectivity asymptotes
        catchability = .01,
        # overwritten by tune_fleet but can be set manually here
        p_explt = 1
      ),
      "bigeye" = Metier$new(
        critter = fauna$bigeye,
        price = 10,
        sel_form = "logistic",
        sel_start = .1,
        sel_delta = .01,
        catchability = 0,
        p_explt = 1
      )
    ),
    base_effort = prod(resolution),
    resolution = resolution
  ),
  "purseseine" = create_fleet(
    list(
      skipjack = Metier$new(
        critter = fauna$skipjack,
        price = 100,
        sel_form = "logistic",
        sel_start = 0.25,
        sel_delta = .1,
        catchability = .01,
        p_explt = 0.9
      ),
      bigeye = Metier$new(
        critter = fauna$bigeye,
        price = 100,
        sel_form = "logistic",
        sel_start = .25,
        sel_delta = .5,
        catchability = .01,
        p_explt = 1
      )
    ),
    base_effort = prod(resolution), resolution = resolution
  )
)

a <- Sys.time()

fleets <- tune_fleets(fauna, fleets)

Sys.time() - a


# run simulations

# run the simulation using marlin::simmar
a <- Sys.time()

sim3 <- simmar(
  fauna = fauna,
  fleets = fleets,
  years = years
)

Sys.time() - a
# a <- Sys.time()

processed_marlin <- process_marlin(sim = sim3, time_step = time_step, keep_age = TRUE)
# Sys.time() - a

plot_marlin(processed_marlin)

plot_marlin(processed_marlin, plot_var = "c")

plot_marlin(processed_marlin, plot_var = "n", plot_type = "length", fauna = fauna)
```


## Evaluating MPAs

Now let's compare the effect of an MPA on two species: a shark population and a yellowfin tuna population. In one simulation, the sharks and the tuna will share the same nearshore habitat to the "east" of the simulated area. In the second, the tunas life onshore in the eastern habitat, but the sharks live further offshore. In both, we will design an MPA based on the distribution of tunas. 



```{r}
resolution <- c(20, 20) # resolution is in squared patches, so 20 implies a 20X20 system, i.e. 400 patches

seasons <- 1

years <- 50

tune_type <- "depletion"

steps <- years * seasons

yft_diffusion <- 6

yft_depletion <- 0.5

mako_depletion <- 0.4

mako_diffusion <- 5

yft_b0 <- 1000

mako_b0 <- 42

# for now make up some habitat

yft_habitat <- expand_grid(x = 1:resolution[1], y = 1:resolution[2]) %>%
  mutate(
    habitat = .05 * x,
    habitat = habitat / max(habitat) * yft_diffusion
  ) %>%
  pivot_wider(names_from = x, values_from = habitat) %>%
  select(-y) %>%
  as.matrix()


mako_habitat <- expand_grid(x = 1:resolution[1], y = 1:resolution[2]) %>%
  mutate(
    habitat = dnorm(x, resolution, 8),
    habitat = habitat / max(habitat) * mako_diffusion
  ) %>%
  pivot_wider(names_from = x, values_from = habitat) %>%
  select(-y) %>%
  as.matrix()


# create a fauna object, which is a list of lists
fauna <-
  list(
    "Yellowfin Tuna" = create_critter(
      scientific_name = "Thunnus albacares",
      habitat = yft_habitat, # pass habitat as lists
      recruit_habitat = yft_habitat,
      adult_diffusion = yft_diffusion, # cells per year
      fished_depletion = yft_depletion, # desired equilibrium depletion with fishing (1 = unfished, 0 = extinct),
      density_dependence = "global_habitat", # recruitment form, where 1 implies local recruitment
      seasons = seasons,
      b0 = yft_b0
    ),
    "Shortfin Mako" = create_critter(
      scientific_name = "Isurus oxyrinchus",
      habitat = list(mako_habitat), # pass habitat as lists
      recruit_habitat = mako_habitat,
      adult_diffusion = mako_diffusion,
      fished_depletion = mako_depletion,
      density_dependence = "global_habitat", # recruitment form, where 1 implies local recruitment
      burn_years = 200,
      seasons = seasons,
      fec_form = "pups",
      pups = 6,
      b0 = mako_b0,
      lorenzen_m = TRUE,
      lorenzen_c = -0.25
    )
  )

fauna$`Shortfin Mako`$plot()
# create a fleets object, which is a list of lists (of lists). Each fleet has one element,
# with lists for each species inside there. Price specifies the price per unit weight of that
# species for that fleet
# sel_form can be one of logistic or dome

fleets <- list("longline" = create_fleet(
  list(
    `Yellowfin Tuna` = Metier$new(
      critter = fauna$`Yellowfin Tuna`,
      price = 100, # price per unit weight
      sel_form = "logistic", # selectivity form, one of logistic or dome
      sel_start = .3, # percentage of length at maturity that selectivity starts
      sel_delta = .1, # additional percentage of sel_start where selectivity asymptotes
      catchability = .01, # overwritten by tune_fleet but can be set manually here
      p_explt = 1
    ),
    `Shortfin Mako` = Metier$new(
      critter = fauna$`Shortfin Mako`,
      price = 0,
      sel_form = "logistic",
      sel_start = .1,
      sel_delta = .01,
      catchability = 0,
      p_explt = 1
    )
  ),
  mpa_response = "stay",
  base_effort = prod(resolution),
  resolution = resolution
))

a <- Sys.time()

fleets <- tune_fleets(fauna, fleets, tune_type = tune_type) # tunes the catchability by fleet to achieve target depletion

Sys.time() - a

# run simulations

a <- Sys.time()

nearshore <- simmar(
  fauna = fauna,
  fleets = fleets,
  years = years
)

Sys.time() - a

proc_nearshore <- process_marlin(nearshore, time_step = fauna[[1]]$time_step)

plot_marlin(proc_nearshore, max_scale = FALSE, plot_var = "b")
plot_marlin(proc_nearshore, max_scale = FALSE, plot_var = "ssb")
```


We will now design and implement an MPA network by specifying a data frame with columns x,y, and mpa denoting the coordinates of MPA patches. 

```{r}
set.seed(42)
# specify some MPA locations
mpa_locations <- expand_grid(x = 1:resolution[1], y = 1:resolution[2]) %>%
  mutate(mpa = x > 15 & y < 15)

mpa_locations %>%
  ggplot(aes(x, y, fill = mpa)) +
  geom_tile() +
  scale_fill_brewer(palette = "Accent", direction = -1, name = "MPA") +
  scale_x_continuous(name = "Lon") +
  scale_y_continuous(name = "Lat")
```

We will now simulate the impacts of those MPAs by passing them to the manager slot. 

```{r}
a <- Sys.time()

nearshore_mpa <- simmar(
  fauna = fauna,
  fleets = fleets,
  years = years,
  manager = list(mpas = list(
    locations = mpa_locations,
    mpa_year = floor(years * .5)
  ))
)

Sys.time() - a

proc_nearshore_mpa <- process_marlin(nearshore_mpa, time_step = fauna[[1]]$time_step)

plot_marlin(proc_nearshore_mpa, max_scale = FALSE)
```


Now though, consider a different scenario. Here the tunas still slightly prefer their same nearshore habitat, but now the shortfin mako population primarily lives farther offshore. We will first simulate that population without the MPA, and then assess the effects of the exact same MPA on this new scenario. 


```{r}
mako_habitat <- expand_grid(x = 1:resolution[1], y = 1:resolution[2]) %>%
  mutate(
    habitat = dnorm(x, .3 * resolution, 8),
    habitat = habitat / max(habitat) * mako_diffusion
  ) %>%
  pivot_wider(names_from = x, values_from = habitat) %>%
  select(-y) %>%
  as.matrix()


# create a fauna object, which is a list of lists

fauna <-
  list(
    "Yellowfin Tuna" = create_critter(
      scientific_name = "Thunnus albacares",
      habitat = yft_habitat, # pass habitat as lists
      recruit_habitat = yft_habitat,
      adult_diffusion = yft_diffusion, # cells per year
      fished_depletion = yft_depletion, # desired equilibrium depletion with fishing (1 = unfished, 0 = extinct),
      density_dependence = "global_habitat", # recruitment form, where 1 implies local recruitment
      seasons = seasons,
      b0 = yft_b0
    ),
    "Shortfin Mako" = create_critter(
      scientific_name = "Isurus oxyrinchus",
      habitat = list(mako_habitat), # pass habitat as lists
      recruit_habitat = mako_habitat,
      adult_diffusion = mako_diffusion,
      fished_depletion = mako_depletion,
      density_dependence = "global_habitat", # recruitment form, where 1 implies local recruitment
      burn_years = 200,
      seasons = seasons,
      fec_form = "pups",
      pups = 6,
      b0 = mako_b0,
      lorenzen_m = TRUE,
      lorenzen_c = -0.25
    )
  )

fauna$`Shortfin Mako`$plot()

fleets <- tune_fleets(fauna, fleets, tune_type = tune_type) # tunes the catchability by fleet to achieve target depletion

# run simulations

# run the simulation using marlin::simmar
a <- Sys.time()

offshore <- simmar(
  fauna = fauna,
  fleets = fleets,
  years = years
)

Sys.time() - a

proc_offshore <- process_marlin(offshore, time_step = fauna[[1]]$time_step)

a <- Sys.time()

offshore_mpa_sim <- simmar(
  fauna = fauna,
  fleets = fleets,
  years = years,
  manager = list(mpas = list(
    locations = mpa_locations,
    mpa_year = floor(years * .5)
  ))
)

Sys.time() - a


proc_offshore_mpa <- process_marlin(offshore_mpa_sim, time_step = fauna[[1]]$time_step)
```



```{r}
plot_marlin(
  `MPA: Sharks Offshore` = proc_offshore_mpa,
  `No MPA: Sharks Nearshore` = proc_nearshore,
  `MPA: Sharks Nearshore` = proc_nearshore_mpa,
  steps_to_plot = NA,
  plot_var = "ssb",
  max_scale = FALSE
)
```

```{r}
plot_marlin(
  `MPA: Sharks Offshore` = proc_offshore_mpa,
  `Sharks Nearshore` = proc_nearshore,
  `Sharks Offshore` = proc_offshore,
  `MPA: Sharks Nearshore` = proc_nearshore_mpa,
  plot_var = "ssb",
  plot_type = "space",
  steps_to_plot = c(years - 1)
) +
  scale_fill_viridis_c(name = "Spawning Biomass", guide = guide_colorbar(barwidth = unit(13, "lines"), frame.colour = "black")) +
  theme(
    axis.text.x = element_blank(),
    axis.text.y = element_blank(),
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    axis.ticks = element_blank(),
    strip.text = element_text(size = 9)
  )
```


## Defacto MPAs through bycatch penalties

We can also run a case where prices for shortfin mako are negative creating *de facto* MPAs, by imposing a severe negative price on shortfin mako that causes the fishing fleet to avoid that area. 


```{r}
years <- 100

tune_type <- "explt"

# make up some habitat

yft_habitat <- expand_grid(x = 1:resolution[1], y = 1:resolution[2]) %>%
  mutate(
    habitat = .05 * x,
    habitat = habitat / max(habitat) * yft_diffusion
  ) %>%
  pivot_wider(names_from = x, values_from = habitat) %>%
  select(-y) %>%
  as.matrix()


mako_habitat <- expand_grid(x = 1:resolution[1], y = 1:resolution[2]) %>%
  mutate(
    habitat = x > 12 & y > 12,
    habitat = habitat / max(habitat) * mako_diffusion
  ) %>%
  pivot_wider(names_from = x, values_from = habitat) %>%
  select(-y) %>%
  as.matrix()


# create a fauna object, which is a list of lists

fauna <-
  list(
    "Yellowfin Tuna" = create_critter(
      scientific_name = "Thunnus albacares",
      habitat = list(yft_habitat),
      recruit_habitat = yft_habitat,
      adult_diffusion = yft_diffusion,
      fished_depletion = .4,
      density_dependence = "local_habitat", # recruitment form, where 1 implies local recruitment
      seasons = seasons,
      init_explt = 0.12,
      explt_type = "f"
    ),
    "Shortfin Mako" = create_critter(
      scientific_name = "Isurus oxyrinchus",
      habitat = list(mako_habitat),
      recruit_habitat = mako_habitat,
      adult_diffusion = mako_diffusion,
      fished_depletion = .3,
      density_dependence = "local_habitat", # recruitment form, where 1 implies local recruitment
      burn_years = 200,
      seasons = seasons,
      init_explt = 0.1,
      explt_type = "f"
    )
  )

# create a fleets object, accounting a negative price to shortfin makos

fleets <- list("longline" = create_fleet(
  list(
    `Yellowfin Tuna` = Metier$new(
      critter = fauna$`Yellowfin Tuna`,
      price = 100, # price per unit weight
      sel_form = "logistic", # selectivity form, one of logistic or dome
      sel_start = .3, # percentage of length at maturity that selectivity starts
      sel_delta = .1, # additional percentage of sel_start where selectivity asymptotes
      catchability = .01, # overwritten by tune_fleet but can be set manually here
      p_explt = 1
    ),
    `Shortfin Mako` = Metier$new(
      critter = fauna$`Shortfin Mako`,
      price = -20000,
      sel_form = "logistic",
      sel_start = .1,
      sel_delta = .01,
      catchability = 0,
      p_explt = 1
    )
  ),
  mpa_response = "stay",
  base_effort = prod(resolution),
  resolution = resolution
))

a <- Sys.time()

fleets <- tune_fleets(fauna, fleets, tune_type = tune_type) # tunes the catchability by fleet to achieve target depletion

Sys.time() - a

# run simulations

# run the simulation using marlin::simmar
negative_prices <- simmar(
  fauna = fauna,
  fleets = fleets,
  years = years
)

proc_negative_prices <- process_marlin(negative_prices, time_step = fauna[[1]]$time_step)
```


```{r}
plot_marlin(
  `De-Facto MPA` = proc_negative_prices,
  plot_var = "ssb",
  plot_type = "space"
)
```



## Repo Naviation

The core wrapper function is located in R/simmar.R. This function keeps track of each of the populations and fleets. 

The actual population models are found in src/fish_model.cpp. Additional modules will be put in there as they are developed
