---
title: "Setting flexible patch areas"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{flexible_patch_area}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE
)
```

One new feature of `marlin` is allowing for a flexible `patch_area()` parameter, in which users can supply a single value (if all grid cells are the same size), or a matrix or vector of values denoting different grid cell areas. This is particularly useful for modeling species with large habitat ranges with a particular focus on a small subset of the total modeled area. 

```{r echo=FALSE, fig.cap = "The area of interest (red) has a finer resolution than the larger modeled area"}
library(tidyverse)
library(sf)

small_grid <- st_make_grid(st_sfc(st_polygon(list(rbind(c(0,0), c(1,0), c(1,1), c(0,0))))), cellsize = 0.1)
big_grid <- st_make_grid(st_sfc(st_polygon(list(rbind(c(-.5,-.5), c(1.5,-0.5), c(1.5,1.5), c(-0.5,-0.5))))), cellsize = 0.5)

ggplot() + 
  geom_sf(data = big_grid, mapping = aes(fill = "Species stock assessment area")) + 
  geom_sf(data = small_grid, mapping = aes(fill = "Area of interest")) + 
  scale_fill_manual(values = c("Area of interest" = "red", 
                               "Species stock assessment area" = "white")) + 
  theme_void() + 
  labs(fill = NULL) + 
  theme(legend.position = "bottom")
```

```{r echo=FALSE, fig.cap = "Patch areas represented by the area of interest and the species stock assessment area."}
grid <- st_make_grid(st_sfc(st_polygon(list(rbind(c(-.5,-.5), c(1.5,-.5), c(1.5,1.5), c(-.5,-.5))))), 
                     cellsize = 0.1) %>% 
  st_as_sf() %>% 
  mutate(area = c(rep(50*50, 20*5), rep(c(rep(50*50, 5), rep(10*10, 10), rep(50*50, 5)), 10), 
                  rep(50*50, 20*5)))

ggplot() + 
  geom_sf(data = grid, mapping = aes(fill = factor(area), color = factor(area))) + 
  geom_sf(data = big_grid, fill = NA, color = "white") +
  geom_sf(data = small_grid, fill = NA, color = "white") +
  scale_color_manual(values = c("100" = "goldenrod", "2500" = "firebrick"), 
                     guide = "none") +
  scale_fill_manual(values = c("100" = "goldenrod", "2500" = "firebrick"), 
                    labels = c("100" = "100 sq. km", "2500" = "2,500 sq. km")) +
  theme_void() + 
  labs(fill = NULL) + 
  theme(legend.position = "bottom")
```

We begin by using the classic example, where patch area is the same.

```{r}
# Load required libraries
# library(marlin)
devtools::load_all()
library(tidyverse)

# Set marlin parameters
resolution <- c(20,20) # resolution is in squared patches, so 20 implies a 20X20 system, i.e. 400 patches
years <- 25
patch_area <- 1000 # in square km
 
# Produce patch dataframe
patches <- expand_grid(
  x = seq(1, resolution[1]),
  y = seq(1, resolution[2])
) |> 
  mutate(
    patch_id = row_number(),
    area = patch_area
  )

# Generate habitat matrices
recruit_habitat <- sim_habitat("bigeye",kp = .001, resolution = resolution, patch_area = patch_area, output = "list")
adult_habitat <- sim_habitat("bigeye",kp = .001, resolution = resolution, patch_area = patch_area, output = "list")

# Build fauna object
fauna <-
  list(
    "bigeye" = create_critter(
      common_name = "bigeye tuna",
      adult_home_range = 10,
      recruit_home_range = 4,
      habitat = adult_habitat$critter_distributions$bigeye,
      recruit_habitat = recruit_habitat$critter_distributions$bigeye,
      density_dependence = "local_habitat",
      resolution = resolution,
      patch_area = patch_area,
      steepness = 0.6,
      ssb0 = 1000
    )
  )

# Plot movement for fauna 
fauna$bigeye$plot_movement()

# Calculate other patch metrics
patch_things <- patches # save the original output
patches$r0 <- fauna$bigeye$r0s # add the number of recruits each patch can support under unfished conditions
patches$ssb0 <- fauna$bigeye$ssb0_p # add spawning stock biomass by patch

# Plot the recruits/ssb0 starting conditions over space
patches |> 
  pivot_longer(cols = c(r0, ssb0), names_to = "variable", values_to = "value") |> 
  group_by(variable) |>
  mutate(svalue = value / max(value)) |>
  ggplot(aes(x = x, y = y, fill = svalue)) +
  geom_tile() +
  facet_wrap(~variable, scales = "free") +
  scale_fill_viridis_c() +
  theme_minimal()
```

We now adjust patch area to represent two different spatial resolutions: 

```{r}
# Set marlin parameters
resolution <- c(20,20) # resolution is in squared patches, so 20 implies a 20X20 system, i.e. 400 patches
years <- 25
patch_area <- matrix(c(rep(50*50, 20*5), rep(c(rep(50*50, 5), rep(10*10, 10), rep(50*50, 5)), 10),
                  rep(50*50, 20*5)), ncol = resolution[1]) # in square km

# Produce patch dataframe
patches <- expand_grid(
  x = seq(1, resolution[1]),
  y = seq(1, resolution[2])
) |>
  mutate(
    patch_id = row_number()) |> 
  left_join(patch_area %>% 
      as.data.frame() %>%
      dplyr::mutate(y = 1:nrow(.)) %>%
      tidyr::pivot_longer(
        -y,
        names_to = "x",
        values_to = "rec_habitat",
        names_prefix = "V",
        names_transform = list(x = as.integer)
        ) 
  )

# Generate habitat matrices
recruit_habitat <- sim_habitat("bigeye",kp = .001, resolution = resolution, patch_area = patch_area, output = "list")
adult_habitat <- sim_habitat("bigeye",kp = .001, resolution = resolution, patch_area = patch_area, output = "list")

# Build fauna object
fauna <-
  list(
    "bigeye" = create_critter(
      common_name = "bigeye tuna",
      adult_home_range = 10,
      recruit_home_range = 4,
      habitat = adult_habitat$critter_distributions$bigeye,
      recruit_habitat = recruit_habitat$critter_distributions$bigeye,
      density_dependence = "local_habitat",
      resolution = resolution,
      patch_area = patch_area,
      steepness = 0.6,
      ssb0 = 1000
    )
  )

# Plot movement for fauna
fauna$bigeye$plot_movement()

# Calculate other patch metrics
patch_things <- patches # save the original output
patches$r0 <- fauna$bigeye$r0s # add the number of recruits each patch can support under unfished conditions
patches$ssb0 <- fauna$bigeye$ssb0_p # add spawning stock biomass by patch

# Plot the recruits/ssb0 starting conditions over space
patches |>
  pivot_longer(cols = c(r0, ssb0), names_to = "variable", values_to = "value") |>
  group_by(variable) |>
  mutate(svalue = value / max(value)) |>
  ggplot(aes(x = x, y = y, fill = svalue)) +
  geom_tile() +
  facet_wrap(~variable, scales = "free") +
  scale_fill_viridis_c() +
  theme_minimal()
```
