---
title: "Getting Started with marlin"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{getting-started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 5
)
set.seed(42)
```

`marlin` simulates age-structured marine populations and fishing fleets across a two-dimensional seascape. You can use it to explore questions like "how does displaced fishing effort change the outcomes of an MPA?" or "how does fleet behavior shape spatial depletion patterns?" — questions where the structure of the dynamics matters more than precise parameter values.

This vignette walks through the basics: creating species, building fleets, running simulations, and examining the results. Each section introduces one layer of complexity. For deeper dives into specific features, see the other vignettes listed at the end.

```{r setup}
library(marlin)
library(ggplot2)
library(dplyr)
library(tidyr)

theme_set(marlin::theme_marlin(base_size = 14))
```

## The Three Ingredients

Every `marlin` simulation needs three things:

1. **`fauna`** — a named list of species (created by `create_critter`)
2. **`fleets`** — a named list of fishing fleets (created by `create_fleet`)
3. **`simmar()`** — the function that runs the simulation

Optionally, you can also pass a **`manager`** list to `simmar` to add MPAs, quotas, or effort caps, but let's start without one.

## A Minimal Example: One Species, One Fleet

### Creating a species

The `create_critter` function builds an age-structured population. If you supply a `scientific_name` (or `common_name`), `marlin` will look up life history parameters from the [FishLife](https://github.com/James-Thorson-NOAA/FishLife) database. You can override any parameter manually.

A few key arguments to know:

- `resolution`: the size of the spatial grid (e.g., `c(10, 10)` for 100 patches)
- `seasons`: number of time steps per year (4 = quarterly)
- `fished_depletion`: the equilibrium biomass ratio (B/B0) you want when fishing is active. This is a *target* that `tune_fleets` will calibrate to.
- `adult_home_range`: distance (in KM) that 95% of adults move from a source patch per year — controls how mobile the species is
- `density_dependence`: how recruitment relates to space — `"global_habitat"` means recruits settle across the whole domain weighted by habitat, `"local_habitat"` means they stay near where they were spawned

```{r one-species}
resolution <- c(10, 10)
years <- 20
seasons <- 4
time_step <- 1 / seasons

fauna <- list(
  "bigeye" = create_critter(
    common_name = "bigeye tuna",
    adult_home_range = 5,
    recruit_home_range = 10,
    density_dependence = "local_habitat",
    seasons = seasons,
    fished_depletion = 0.4,
    resolution = resolution,
    steepness = 0.6,
    ssb0 = 1000
  )
)
```

The critter object has built-in plots for inspecting life history and movement:

```{r critter-plots}
fauna$bigeye$plot()

fauna$bigeye$plot_movement()
```

### Creating a fleet

Fleets are built with `create_fleet`. Each fleet contains one or more **metiers** — one per species it interacts with — specifying the price, selectivity, catchability, and share of exploitation (`p_explt`).

The most important fleet-level settings are:

- `fleet_model`: `"constant_effort"` (total effort fixed) or `"open_access"` (effort responds to profits)
- `spatial_allocation`: how effort is distributed across patches — `"rpue"` (chase high catch rates) is the default
- `base_effort`: total effort units at the start

```{r one-fleet}
fleets <- list(
  "longline" = create_fleet(
    list(
      "bigeye" = Metier$new(
        critter = fauna$bigeye,
        price = 10,
        sel_form = "logistic",
        sel_start = 1,
        sel_delta = 0.01,
        catchability = 0,
        p_explt = 1
      )
    ),
    base_effort = prod(resolution),
    resolution = resolution,
    fleet_model = "constant_effort"
  )
)
```

### Tuning

You'll notice we set `catchability = 0` and `fished_depletion = 0.4` above. That's because we use `tune_fleets` to find the catchability that produces our target depletion level. This is almost always easier than trying to guess catchability directly.

```{r tune}
fleets <- tune_fleets(fauna, fleets, tune_type = "depletion")
```

### Running the simulation

Pass fauna, fleets, and a number of years to `simmar`:

```{r run-simple}
sim <- simmar(
  fauna = fauna,
  fleets = fleets,
  years = years
)
```

### Examining results

`process_marlin` tidies the raw output into data frames. `plot_marlin` provides quick visualizations.

```{r results-simple}
proc <- process_marlin(sim, time_step = time_step)

# Spawning stock biomass over time
plot_marlin(proc, plot_var = "ssb")

# Catch over time
plot_marlin(proc, plot_var = "c", max_scale = FALSE)

# Spatial distribution of SSB in the final time step
plot_marlin(proc, plot_var = "ssb", plot_type = "space",
            steps_to_plot = max(proc$fauna$step))
```

The output of `process_marlin` is a list with two data frames. `proc$fauna` has population-level results (biomass, numbers, SSB, catch by patch, age, and time step), and `proc$fleets` has fleet-level results (catch, revenue, effort, CPUE by patch, fleet, and time step). You can use these directly for custom analyses:

```{r custom-plot}
proc$fleets %>%
  group_by(step, fleet) %>%
  summarise(catch = sum(catch), .groups = "drop") %>%
  ggplot(aes(step * time_step, catch)) +
  geom_line() +
  labs(x = "Year", y = "Total Catch", title = "Fleet-level catch over time")
```

## Scaling Up: Two Species, Two Fleets

Real systems have multiple species and multiple fleets interacting in space. Let's build a more realistic example with spatially heterogeneous habitat.

### Generating habitat with `sim_habitat`

`sim_habitat` generates spatially correlated habitat maps for multiple species simultaneously. The `kp` parameter controls smoothness (lower = smoother), and you can supply a correlation matrix to control whether species prefer similar or different areas.

```{r habitat}
# Species prefer different areas (negative correlation)
critter_cors <- matrix(c(1, -0.5,
                         -0.5, 1), nrow = 2)

habitats <- sim_habitat(
  critters = c("bigeye", "skipjack"),
  kp = 0.3,
  critter_correlations = critter_cors,
  resolution = resolution,
  patch_area = 1,
  output = "list"
)

bigeye_habitat <- habitats$critter_distributions$bigeye
skipjack_habitat <- habitats$critter_distributions$skipjack
```

### Building the fauna

Now we create two species using these habitats:

```{r two-fauna}
fauna <- list(
  "bigeye" = create_critter(
    common_name = "bigeye tuna",
    habitat = list(bigeye_habitat),
    season_blocks = list(1:seasons),
    adult_home_range = 5,
    recruit_home_range = 10,
    density_dependence = "local_habitat",
    seasons = seasons,
    fished_depletion = 0.4,
    resolution = resolution,
    steepness = 0.6,
    ssb0 = 1000
  ),
  "skipjack" = create_critter(
    scientific_name = "Katsuwonus pelamis",
    habitat = list(skipjack_habitat),
    season_blocks = list(1:seasons),
    adult_home_range = 3,
    recruit_home_range = 8,
    density_dependence = "local_habitat",
    seasons = seasons,
    fished_depletion = 0.6,
    resolution = resolution,
    steepness = 0.7,
    ssb0 = 800
  )
)
```

### Building the fleets

Each fleet needs a metier for every species in the fauna, even if a fleet doesn't target a species (set `p_explt = 0` for bycatch-only or non-interaction). The `p_explt` values are relative: if longline has `p_explt = 2` for bigeye and purse seine has `p_explt = 1`, the longline accounts for 2/3 of bigeye fishing mortality.

```{r two-fleets}
fleets <- list(
  "longline" = create_fleet(
    list(
      "bigeye" = Metier$new(
        critter = fauna$bigeye,
        price = 10,
        sel_form = "logistic",
        sel_start = 1,
        sel_delta = 0.01,
        catchability = 0,
        p_explt = 2
      ),
      "skipjack" = Metier$new(
        critter = fauna$skipjack,
        price = 5,
        sel_form = "logistic",
        sel_start = 0.8,
        sel_delta = 0.1,
        catchability = 0,
        p_explt = 1
      )
    ),
    base_effort = prod(resolution),
    resolution = resolution,
    fleet_model = "constant_effort"
  ),
  "purseseine" = create_fleet(
    list(
      "bigeye" = Metier$new(
        critter = fauna$bigeye,
        price = 10,
        sel_form = "logistic",
        sel_start = 0.5,
        sel_delta = 0.5,
        catchability = 0,
        p_explt = 1
      ),
      "skipjack" = Metier$new(
        critter = fauna$skipjack,
        price = 8,
        sel_form = "logistic",
        sel_start = 0.3,
        sel_delta = 0.1,
        catchability = 0,
        p_explt = 2
      )
    ),
    base_effort = prod(resolution),
    resolution = resolution,
    fleet_model = "constant_effort"
  )
)

fleets <- tune_fleets(fauna, fleets, tune_type = "depletion")
```

### Running and examining

```{r run-multi}
sim_multi <- simmar(fauna = fauna, fleets = fleets, years = years)

proc_multi <- process_marlin(sim_multi, time_step = time_step)

plot_marlin(proc_multi, plot_var = "ssb")

plot_marlin(proc_multi, plot_var = "c", max_scale = FALSE)

plot_marlin(proc_multi, plot_var = "ssb", plot_type = "space",
            steps_to_plot = max(proc_multi$fauna$step))
```

## Adding an MPA

One of the main use cases for `marlin` is evaluating the impacts of marine protected areas. You specify MPA locations as a data frame with `x`, `y`, and `mpa` (logical) columns, and pass it to `simmar` via the `manager` argument. The `mpa_year` parameter sets when the MPA is established.

```{r mpa-setup}
# Protect a block of patches in the upper-right corner
mpa_locations <- expand_grid(x = 1:resolution[1], y = 1:resolution[2]) %>%
  mutate(mpa = x > 7 & y > 7)

# Visualize the MPA network
ggplot(mpa_locations, aes(x, y, fill = mpa)) +
  geom_tile() +
  scale_fill_brewer(palette = "Accent", direction = -1, name = "MPA") +
  coord_equal() +
  labs(title = "MPA Network")
```

```{r mpa-run}
sim_mpa <- simmar(
  fauna = fauna,
  fleets = fleets,
  years = years,
  manager = list(
    mpas = list(
      locations = mpa_locations,
      mpa_year = 10
    )
  )
)

proc_mpa <- process_marlin(sim_mpa, time_step = time_step)
```

`plot_marlin` can overlay multiple simulation results for comparison. Pass named arguments to get labeled panels:

```{r mpa-compare}
plot_marlin(
  "No MPA" = proc_multi,
  "With MPA" = proc_mpa,
  plot_var = "ssb"
)

plot_marlin(
  "No MPA" = proc_multi,
  "With MPA" = proc_mpa,
  plot_var = "c",
  max_scale = FALSE
)
```

You can also compare spatial patterns:

```{r mpa-spatial}
plot_marlin(
  "No MPA" = proc_multi,
  "With MPA" = proc_mpa,
  plot_var = "ssb",
  plot_type = "space",
  steps_to_plot = max(proc_mpa$fauna$step)
)
```

## Next Steps

This vignette covered the basics. `marlin` has a lot more to offer — here are the vignettes to explore next depending on what you need:

**Fleet behavior:**

- [Spatial Allocation Strategies](spatial-allocation.html) — how `rpue`, `ppue`, `profit`, `marginal_profit`, and other allocation rules shape where fleets fish
- [Setting and Using Distance from Port](port-distance.html) — adding travel costs and port locations
- [Manage Catch and Effort of Fleets](fleet-management.html) — open access dynamics, quotas, effort caps, and manual effort
- [Manual Fishing Grounds and Effort Allocation](manual_fleets.html) — controlling exactly where fleets can fish

**Species biology:**

- [Setting & Understanding Movement Rates](movement-rates.html) — how home range and diffusion work
- [Set Dynamic Habitats](dynamic-habitat.html) — seasonal habitat shifts and climate-driven range changes
- [Setting Selectivity](Setting-Selectivity.html) — logistic, dome, and manual selectivity options
- [Working with Recruitment Deviates](Working-with-Recruitment-Deviates.html) — stochastic recruitment
- [Invertebrates](Invertebrates.html) — non-fish species

**Analysis and design:**

- [Management Strategy Evaluation](MSE.html) — running `simmar` iteratively with adaptive management
- [Measure MPA Gradients](measure_gradients.html) — quantifying spillover and edge effects
