---
title: "Measure MPA Gradients"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{measure_gradients}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

MPA performance is often measured by examining gradients of various outcomes, such as catch-per-unit-effort (CPUE) inside vs. various distances outside the protected area. 
Here is an example of generating these kinds of measurements using `marlin`


```{r, message=FALSE, warning=FALSE}

library(marlin)

library(tidyverse)

library(patchwork)

theme_set(marlin::theme_marlin())

resolution <- c(5,20) # resolution is in squared patches, so 20 implies a 20X20 system, i.e. 400 patches 

patch_area <- 10

seasons <- 2

years <- 20

tune_type <- "depletion"

steps <- years * seasons

yft_diffusion <- 5
  
yft_depletion <- .2

rec_factor <- 10

mako_depletion <- 0.1

mako_diffusion <- 100

# for now make up some habitat

yft_habitat <- expand_grid(x = 1:resolution[1], y = 1:resolution[2]) %>%
  mutate(habitat =  1,
         habitat = habitat / max(habitat) * yft_diffusion) %>% 
  pivot_wider(names_from = y, values_from = habitat) %>% 
  select(-x) %>% 
  as.matrix()
 

mako_habitat <- expand_grid(x = 1:resolution[1], y = 1:resolution[2]) %>%
  mutate(habitat =  1,
         habitat = habitat / max(habitat) * mako_diffusion) %>% 
  pivot_wider(names_from = y, values_from = habitat) %>% 
  select(-x) %>% 
  as.matrix()


# create a fauna object, which is a list of lists
fauna <- 
  list(
    "Yellowfin Tuna" = create_critter(
      scientific_name = "Thunnus albacares",
      habitat = yft_habitat, # pass habitat as lists
      recruit_habitat = yft_habitat,
      adult_diffusion = yft_diffusion, # cells per year
      recruit_diffusion = rec_factor * yft_diffusion,
      fished_depletion = yft_depletion, # desired equilibrium depletion with fishing (1 = unfished, 0 = extinct),
      density_dependence = "pre_dispersal", # recruitment form, where 1 implies local recruitment
      seasons = seasons,
      ssb0 = 5000,
      ),
    "Shortfin Mako" = create_critter(
      scientific_name = "Isurus oxyrinchus",
      habitat = list(mako_habitat), # pass habitat as lists
      recruit_habitat = mako_habitat,
      adult_diffusion = mako_diffusion,
      recruit_diffusion = rec_factor * .1,
      fished_depletion = mako_depletion,
      density_dependence = "local_habitat", # recruitment form, where 1 implies local recruitment
      burn_years = 200,
      ssb0 = 100,
      seasons = seasons,
      fec_form = "pups",
      pups = 2,
      lorenzen_m = FALSE
    )
  )


fleets <- list("longline" = create_fleet(list(
  `Yellowfin Tuna` = Metier$new(
    critter = fauna$`Yellowfin Tuna`,
    price = 100, # price per unit weight
    sel_form = "logistic", # selectivity form, one of logistic or dome
    sel_start = .3, # percentage of length at maturity that selectivity starts
    sel_delta = .1, # additional percentage of sel_start where selectivity asymptotes
    catchability = .01, # overwritten by tune_fleet but can be set manually here
    p_explt = 1
    ),
  `Shortfin Mako` = Metier$new(
    critter = fauna$`Shortfin Mako`,
    price = 10,
    sel_form = "logistic",
    sel_start = .1,
    sel_delta = .01,
    catchability = 0,
    p_explt = 1
  )),
  mpa_response = "stay",
  base_effort = prod(resolution),
  spatial_allocation = "rpue",
  resolution = resolution
))

a <- Sys.time()

fleets <- tune_fleets(fauna, fleets, tune_type = tune_type) # tunes the catchability by fleet to achieve target depletion

Sys.time() - a

# run simulations

a <- Sys.time()

spillover_sim <- simmar(fauna = fauna,
                  fleets = fleets,
                  years = years)

Sys.time() - a
  
proc_spillover <- process_marlin(spillover_sim, time_step =  fauna[[1]]$time_step)

plot_marlin(proc_spillover, max_scale = FALSE)

mpa_locations <- expand_grid(x = 1:resolution[1], y = 1:resolution[2]) |> 
  arrange(x) |> 
  mutate(patch_name = paste(x,y, sep = "_"))

mpas <- mpa_locations$patch_name[1:round(nrow(mpa_locations) * .3)]

# mpas <- sample(mpa_locations$patch_name, round(nrow(mpa_locations) * 0.2), replace = FALSE, prob = mpa_locations$x)

mpa_locations$mpa <- mpa_locations$patch_name %in% mpas
  
a <- Sys.time()

mpa_spillover <- simmar(
  fauna = fauna,
  fleets = fleets,
  years = years,
  manager = list(mpas = list(locations = mpa_locations,
              mpa_year = floor(years * .5)))
)

Sys.time() - a

proc_mpa_spillover <- process_marlin(mpa_spillover, time_step =  fauna[[1]]$time_step)

plot_marlin(proc_mpa_spillover, max_scale = FALSE) + 
  labs(y = "Spawning Stock Biomass")

plot_marlin(proc_mpa_spillover, max_scale = FALSE, plot_var = "c") + 
  labs(y = "Catch")



```

We can use `marlin::get_distance_to_mpas` to measure the euclidian distance between the centroid of each cell and the nearest MPA cell centroid. 

```{r fig-mpa-dist, fig.cap="Distance of each cell to the nearest MPA edge."}

mpa_distances <- get_distance_to_mpas(mpa_locations = mpa_locations, resolution = resolution, patch_area = patch_area)

  mpa_distances |>
    ggplot(aes(x,y,fill = distance_to_mpa_edge)) +
    geom_tile() +
    geom_tile(aes(x,y, color = mpa), size = 1.5) +
    scale_fill_gradient2(low = "darkblue", high = "green", mid = "white", midpoint = 0)
```

```{r fig-dist2, fig.cap="MPA gravity per cell."}

  mpa_distances |>
    ggplot(aes(x,y,fill = total_mpa_distance)) +
    geom_tile() +
    geom_tile(aes(x,y, color = mpa)) + 
  scale_fill_viridis_c(name = "MPA gravity",option = "magma", direction = -1)

```

We can then plot the gradients of various metrics in space and as a function of MPA distance. 

We can see first off that the model generates "fishing the line" behavior where the fishing fleet is concentrated more around the borders of the protected area. 

```{r fig-line, fig.cap="Distribution of fishing effort in space post-MPA."}

conservation_outcomes <- proc_mpa_spillover$fauna 
  
fishery_outcomes <- proc_mpa_spillover$fleets 


a = fishery_outcomes |> 
  filter(step == max(step), effort > 0) |> 
    left_join(mpa_distances, by = c("x","y","patch")) |> 
  ggplot(aes(distance_to_mpa_edge, effort, color = factor(step))) + 
  geom_point() +
  facet_wrap(~critter, scales = "free_y")


b = fishery_outcomes |>
  mutate(effort = if_else(effort == 0, NA, effort)) |> 
  filter(step == max(step)) |>
  ggplot(aes(x, y, fill = effort)) +
  geom_tile()

a / b
```

The complex interactions of the life history and fishing fleet strategies mean that this "fishing the line" behavior results in a linear biomass gradient with distance from the MPA for the tuna, but actually a reverse trend for the shark where biomass is depressed right by the border due to fishing the line, and then increases with distance. 

Note that we see clear "fishing the line" and spillover gradients even though this MPA has had almost no impact on total biomass or

```{r fig-spillover, fig.cap="Biomass of each species as a function of distance from MPA border. Negative distance means inside the MPA."}

conservation_outcomes |>
  filter(step == max(step)) |>
  group_by(step, x, y, patch, critter) |>
  summarise(biomass = sum(b)) |>
  left_join(mpa_distances, by = c("x", "y", "patch")) |>
  ggplot(aes(distance_to_mpa_edge, biomass, color = factor(step))) +
  geom_vline(xintercept = 0) +
  geom_jitter() +
  facet_wrap( ~ critter, scales = "free_y")


```
