---
title: "Measure MPA Gradients"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{measure_gradients}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

MPA performance is often attempted to be measured by empirical indicators, such as the biomass inside the reserve relative to the biomass in a selected reference site outside the reserve. This vignette shows an example of how to simulate those processes using `marlin`. 


We'll first simulate a system with some heterogeneous habitat.  

```{r, message=FALSE, warning=FALSE}

library(marlin)

library(tidyverse)

library(patchwork)

library(rstanarm)

library(ggdist)

library(bayesplot)

theme_set(marlin::theme_marlin())

resolution <- c(10,10) # resolution is in squared patches, so 20 implies a 20X20 system, i.e. 400 patches 

patch_area <- 10

seasons <- 2

years <- 20

tune_type <- "depletion"

steps <- years * seasons

yft_diffusion <- 2
  
yft_depletion <- .2

rec_factor <- 10

mako_depletion <- 0.3

mako_diffusion <- 100

# for now make up some habitat

critters <- c("yellowfin", "mako")

critter_correlation <- -0.4

corr_mat <- matrix( c(1,critter_correlation,critter_correlation,1) , nrow=2 ) # correlation matrix
    

habitats <- sim_habitat(
  critters,
  kp = 0.1,
  critter_correlations = corr_mat,
  resolution = resolution,
  patch_area = patch_area
)

    
critter_habitats <- habitats$critter_distributions |>
  group_by(critter) |>
  nest() |>
  mutate(habitat = map(
    data,
    \(x) x |> select(-patch) |> pivot_wider(names_from = y, values_from = habitat) |>
      select(-x) %>%
      as.matrix()
  ))
  


```


```{r, fig.cap="Simulated Species Habitat"}
habitats$critter_distributions |> 
  ggplot(aes(x,y,fill = habitat)) + 
  geom_tile() + 
  facet_wrap(~critter) + 
  scale_fill_viridis_c()
```


```{r}
# create a fauna object, which is a list of lists

yft_habitat <- critter_habitats$habitat[[which(critter_habitats$critter == "yellowfin")]]

mako_habitat <- critter_habitats$habitat[[which(critter_habitats$critter == "mako")]]

fauna <- 
  list(
    "yellowfin" = create_critter(
      scientific_name = "Thunnus albacares",
      habitat = yft_habitat, # pass habitat as lists
      recruit_habitat = yft_habitat,
      adult_diffusion = yft_diffusion, # cells per year
      recruit_diffusion = rec_factor * yft_diffusion,
      fished_depletion = yft_depletion, # desired equilibrium depletion with fishing (1 = unfished, 0 = extinct),
      density_dependence = "pre_dispersal", # recruitment form, where 1 implies local recruitment
      seasons = seasons,
      ssb0 = 5000,
      ),
    "mako" = create_critter(
      scientific_name = "Isurus oxyrinchus",
    habitat = mako_habitat, 
      recruit_habitat = mako_habitat,
      adult_diffusion = mako_diffusion,
      recruit_diffusion = rec_factor * .1,
      fished_depletion = mako_depletion,
      density_dependence = "local_habitat", # recruitment form, where 1 implies local recruitment
      burn_years = 200,
      ssb0 = 100,
      seasons = seasons,
      fec_form = "pups",
      pups = 2,
      lorenzen_m = FALSE
    )
  )


fleets <- list("longline" = create_fleet(list(
  `yellowfin` = Metier$new(
    critter = fauna$`yellowfin`,
    price = 100, # price per unit weight
    sel_form = "logistic", # selectivity form, one of logistic or dome
    sel_start = .3, # percentage of length at maturity that selectivity starts
    sel_delta = .1, # additional percentage of sel_start where selectivity asymptotes
    catchability = .01, # overwritten by tune_fleet but can be set manually here
    p_explt = 1
    ),
  `mako` = Metier$new(
    critter = fauna$`mako`,
    price = 10,
    sel_form = "logistic",
    sel_start = .1,
    sel_delta = .01,
    catchability = 0,
    p_explt = 1
  )),
  mpa_response = "stay",
  base_effort = 1000 * prod(resolution),
  spatial_allocation = "rpue",
  resolution = resolution
))

a <- Sys.time()

fleets <- tune_fleets(fauna, fleets, tune_type = tune_type) # tunes the catchability by fleet to achieve target depletion

Sys.time() - a

# run simulations

a <- Sys.time()

spillover_sim <- simmar(fauna = fauna,
                  fleets = fleets,
                  years = years)

Sys.time() - a


  
proc_spillover <- process_marlin(spillover_sim, time_step =  fauna[[1]]$time_step, keep_age = TRUE)

plot_marlin(
  proc_spillover,
  max_scale = TRUE,
  plot_var = "b",
  plot_type = "space"
)

plot_marlin(proc_spillover, max_scale = FALSE)

mpa_locations <- expand_grid(x = 1:resolution[1], y = 1:resolution[2]) |> 
  arrange(x) |> 
  mutate(patch_name = paste(x,y, sep = "_"))

mpas <- mpa_locations$patch_name[1:round(nrow(mpa_locations) * .3)]

# mpas <- sample(mpa_locations$patch_name, round(nrow(mpa_locations) * 0.2), replace = FALSE, prob = mpa_locations$x)

mpa_locations$mpa <- mpa_locations$patch_name %in% mpas
  
a <- Sys.time()

mpa_spillover <- simmar(
  fauna = fauna,
  fleets = fleets,
  years = years,
  manager = list(mpas = list(locations = mpa_locations,
              mpa_year = floor(years * .5)))
)

Sys.time() - a

proc_mpa_spillover <- process_marlin(mpa_spillover, time_step =  fauna[[1]]$time_step)


```

We can use `marlin::get_distance_to_mpas` to measure the euclidian distance between the centroid of each cell and the nearest MPA cell centroid. 

```{r fig-mpa-dist, fig.cap="Distance of each cell to the nearest MPA edge."}

mpa_distances <- get_distance_to_mpas(mpa_locations = mpa_locations, resolution = resolution, patch_area = patch_area)

  mpa_distances |>
    ggplot(aes(x,y,fill = distance_to_mpa_edge)) +
    geom_tile() +
    geom_tile(aes(x,y, color = mpa), size = 1.5) +
    scale_fill_gradient2(low = "darkblue", high = "green", mid = "white", midpoint = 0)
```

We can also calculate an "MPA gravity", measured as the sum of all the distances from the current cell to all MPA cells. So, a single MPA patch far from any other MPAs might actually have a lower MPA gravity than a fished patch right next to a very large number of MPA cells. 

```{r fig-dist2, fig.cap="MPA gravity per cell."}

  mpa_distances |>
    ggplot(aes(x,y,fill = -total_mpa_distance)) +
    geom_tile() +
    geom_tile(aes(x,y, color = mpa), size = 1.5) + 
  scale_fill_viridis_c(name = "MPA gravity",option = "magma", direction = -1)

```
### Indicators

We can then use these simulated data to generate a series of indicators of MPA performance


#### Fishing the Line



The complex interactions of the life history and fishing fleet strategies mean that this "fishing the line" behavior results in a linear biomass gradient with distance from the MPA for the tuna, but actually a reverse trend for the shark where biomass is depressed right by the border due to fishing the line, and then increases with distance. 


```{r fig-line, fig.cap="Distribution of fishing effort in space post-MPA."}

conservation_outcomes <- proc_mpa_spillover$fauna 
  
fishery_outcomes <- proc_mpa_spillover$fleets 


a = fishery_outcomes |> 
  filter(step == max(step), effort > 0) |> 
    left_join(mpa_distances, by = c("x","y","patch")) |> 
  ggplot(aes(distance_to_mpa_edge, effort, color = factor(step))) + 
  geom_point() +
  facet_wrap(~critter, scales = "free_y")


b = fishery_outcomes |>
  mutate(effort = if_else(effort == 0, NA, effort)) |> 
  filter(step == max(step)) |>
  ggplot(aes(x, y, fill = effort)) +
  geom_tile()

a / b
```

#### Response Ratios

The most common form of indicator used in the empirical MPA monitoring literature is a "response ratio" *r*, generally measured as some version of 

$$
r = log(\frac{inside}{outside})
$$
Which at lower ratio values is roughly equivalent to the percent difference in the metric in question inside the MPA relative to outside. 

This can be generalized into a regression of the form

$$
log(Y_i) \sim N(\beta_0 + rMPA_i + \pmb{\beta}\pmb{X_i},\sigma)
$$
where $\beta_0$ is the value of outcome *Y* when MPA = 0, MPA is a dummy variable indicating whether observation *i* is inside or outside of an MPA, and $\pmb{\beta}$ is a vector of coefficients for the vector of covariates $\pmb{X}$. 

We can use this equation then to calculate response ratios in various metrics, such as biomass density or mean length of fish. 

```{r}
rr_data <- conservation_outcomes |>
  filter(step == max(step) | year == floor(years * .5)) |>
  group_by(step, x, y, patch, critter) |>
  summarise(biomass = sum(b),
            mean_length = weighted.mean(mean_length, n)) |>
  left_join(mpa_distances, by = c("x", "y", "patch"))




```



```{r fig-bgradient, fig.cap="Biomass of each species as a function of distance from MPA border. Negative distance means inside the MPA."}

rr_data|>
  ggplot(aes(distance_to_mpa_edge, biomass, color = factor(step))) +
  geom_vline(xintercept = 0) +
  geom_jitter() +
  facet_wrap( ~ critter, scales = "free_y")


```


```{r fig-mlgradient, fig.cap="Biomass of each species as a function of distance from MPA border. Negative distance means inside the MPA."}

rr_data|>
  ggplot(aes(distance_to_mpa_edge, mean_length, color = factor(step))) +
  geom_vline(xintercept = 0) +
  geom_jitter() +
  facet_wrap( ~ critter, scales = "free_y") + 
  scale_y_continuous(name = "Mean Length")


```

## Quantifying Gradients


```{r}

model_critter <- "yellowfin"


baseline <- data.frame(ssb0_p = fauna[[model_critter]]$ssb0_p) |>
  mutate(patch = 1:n()) |> 
  mutate(scaled_ssb0_p = as.numeric(scale(ssb0_p)))


mpa_and_ssb <- mpa_distances |>
  left_join(baseline, by = "patch")

survey_indicators = conservation_outcomes |>
  filter(step == max(step), critter == model_critter) |>
  group_by(step, x, y, patch, critter) |>
  summarise(biomass = sum(b),
            mean_length = weighted.mean(mean_length, n)) |>
  left_join(mpa_and_ssb, by = c("x", "y", "patch")) |>
  ungroup() 

response_ratio_model = stan_glm(
  biomass ~ mpa + scaled_ssb0_p - 1,
  data = survey_indicators,
  chains = 1,
  cores = 1
)

response_ratio <- tidybayes::tidy_draws(response_ratio_model) |>
  mutate(response_ratio = mpaTRUE / (mpaFALSE) - 1)


response_ratio |>
  ggplot(aes(response_ratio)) +
  geom_histogram() +
  scale_x_continuous(labels = scales::percent, name = "Response Ratio")

```


### Gradients

```{r}

model_critter <- "mako"


halpern_foo <- function(pars, y, distance_to_mpa, use = "fit") {
  yhat = pars[1] / (1 + pars[2] * exp(pars[3] * distance_to_mpa)) + pars[4]
  
  if (use == "fit") {
    out <- sum(((yhat) - (y)) ^ 2)
  } else {
    out <- yhat
  }
  
}


baseline <- data.frame(ssb0_p = fauna[[model_critter]]$ssb0_p) |>
  mutate(patch = 1:n()) |> 
  mutate(scaled_ssb0_p = as.numeric(scale(ssb0_p)))




survey_indicators = conservation_outcomes |>
  filter(step == max(step), critter == model_critter) |>
  group_by(step, x, y, patch, critter) |>
  summarise(biomass = sum(b),
            mean_length = weighted.mean(mean_length, n)) |>
  left_join(mpa_and_ssb, by = c("x", "y", "patch")) |>
  ungroup() 

fishery_indicators <- fishery_outcomes |>
  filter(step == max(step), critter == model_critter) |>
  group_by(step, x, y, patch, critter, fleet) |>
  summarise(
    mean_length = weighted.mean(mean_length, catch),
    catch = sum(catch),
    effort = sum(effort),
    cpue = sum(catch) / sum(effort)
  ) |>
  left_join(mpa_and_ssb, by = c("x", "y", "patch")) |>
  ungroup() 


test <- nlminb(rep(0,4),lower = c(-Inf,0,-Inf,0),halpern_foo, y = survey_indicators$biomass, distance_to_mpa = survey_indicators$distance_to_mpa_edge, control = list(eval.max = 1000, iter.max = 1000))

out <- halpern_foo(
  test$par,
  y = survey_indicators$biomass,
  distance_to_mpa = survey_indicators$distance_to_mpa_edge,
  use = "predict"
)

plot(survey_indicators$biomass)
lines(out)

spillpars <- test$par

spillpars[4] <- 0

dists <- seq(0,1000)
spillover <- halpern_foo(
  spillpars,
  y = survey_indicators$biomass,
  distance_to_mpa = dists,
  use = "predict"
)

zerotol <- 0.05
spillplot <- data.frame(distance_to_mpa_edge =dists, spillover = spillover) 

spillover_distance <- spillplot |> 
  filter(spillover < (1 + zerotol) * min(spillover)) |> 
  slice(1)

spillplot |> 
  ggplot(aes(distance_to_mpa_edge, spillover)) + 
  geom_point() + 
  geom_vline(xintercept = spillover_distance$distance_to_mpa_edge)


```



Aha, so use the model, then calculate the distance at which the predicted 

```{r}

b_gradient_model = stan_glm(
  biomass ~ distance_to_mpa_edge,
  data = survey_indicators |> filter(!mpa),
  family = Gamma(link = "log"),
  chains = 1,
  cores = 1
)


plot(b_gradient_model)

cpue_gradient_model = stan_glm(
  cpue ~ distance_to_mpa_edge,
  data = fishery_indicators |> filter(!mpa),
  family = Gamma(link = "log"),
  chains = 1,
  cores = 1
)

effort_gradient_model = stan_glm(
  effort ~ distance_to_mpa_edge + scaled_ssb0_p,
  data = fishery_indicators |> filter(!mpa),
  family = Gamma(link = "log"),
  chains = 1,
  cores = 1
)

mcmc_hist(b_gradient_model, pars = "distance_to_mpa_edge")

mcmc_hist(effort_gradient_model, pars = "distance_to_mpa_edge")

predictions <- posterior_predict(b_gradient_model, newdata = survey_indicators |> mutate(scaled_ssb0_p = 0), type = "response") |>
  as.data.frame() |>
  mutate(iteration = 1:n()) |>
  pivot_longer(-iteration, names_to = "patch", values_to = "biomass_hat", 
               names_transform = list(patch = as.integer)) |> 
  left_join(mpa_and_ssb, by = "patch")

survey_indicators |> 
  ggplot(aes(distance_to_mpa_edge, biomass)) + 
  geom_point() + 
  stat_lineribbon(data = predictions, aes(distance_to_mpa_edge, biomass_hat), alpha = 0.5) + 
  scale_fill_brewer() + 
  scale_y_continuous(limits = c(0, NA))


```



