---
title: "MSE"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{MSE}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

`marlin` has a lot of features built into it to simulate dynamics over time. However, we cannot possibly build in every possible use case a researcher might have. 

To that end, and important feature of `marlin` is that the outputs of a `marlin` simulation can be passed as the input to a `marlin` simulation. This means you can run the model for one time step, *insert whatever whacky stuff you want to do*, then run a new step. 

As an example, we'll do a very cartonish version of a "management strategy evaluation", where we 

1. Simulate a fishery for a time step

2. Simulate a perfect stock assessment that tells us *depletion* (biomass / unfished biomass)

3. Applies a harvest control rule that sets a catch quota based on the results of our assessment

4. Run another time step with the supplied quota. 

5. Repeat process

```{r setup}
library(marlin)

library(tidyverse)

theme_set(theme_marlin(base_size = 14))

resolution <- 10 # resolution is in squared patches, so 20 implies a 20X20 system, i.e. 400 patches 

burn_years <- 20

sim_years <- 20

seasons <- 4

time_step <- 1 / seasons

fauna <- 
  list(
    "striped marlin" = create_critter(
      scientific_name = "Kajikia audax",
      adult_diffusion = 10,
      seasons = seasons,
      fished_depletion = 0.2,
      resolution = resolution,
      steepness = 0.6,
      ssb0 = 1000
    )
  )


fleets <- list(
  "longline" = create_fleet(
    list("striped marlin" = Metier$new(
        critter = fauna$`striped marlin`,
        price = 10,
        sel_form = "logistic",
        sel_start = 1,
        sel_delta = .01,
        catchability = 0,
        p_explt = 2
      )
    ),
    base_effort = resolution ^ 2,
    resolution = resolution,
    responsiveness = .5,
    cost_per_unit_effort = 1,
    fleet_model = "open access")
,
"handline" = create_fleet(
  list("striped marlin" = Metier$new(
    critter = fauna$`striped marlin`,
    price = 10,
    sel_form = "logistic",
    sel_start = 1,
    sel_delta = .01,
    catchability = 0,
    p_explt = 1
  )
  ),
  base_effort = resolution ^ 2,
  resolution = resolution,
  fleet_model = "constant effort"
))

fleets <- tune_fleets(fauna, fleets, tune_type = "depletion") 


hcr_cutoff <- 0.2

hcr_target <- 0.5

max_u <- 0.2

depletion <- seq(0,1, by= 0.1)

hcr_slope <- max_u / (hcr_target - hcr_cutoff)

hcr_intercept <- -hcr_slope * hcr_cutoff

hcr <- pmin(max_u,pmax(.01,hcr_slope * depletion + hcr_intercept))

hcr_frame <- data.frame(depletion = depletion, hcr = hcr)

hcr_frame %>% 
  ggplot(aes(depletion, hcr)) + 
  geom_line() + 
  scale_x_continuous(name = "B/B0") + 
  scale_y_continuous(name = "Fishing Mortality Rate")

```


```{r}
mse_sim <- simmar(fauna = fauna,
                  fleets = fleets,
                  years = burn_years)

initial_conditions <- mse_sim[[length(mse_sim)]]

starting_step = as.numeric(last(names(mse_sim)))

for (y in 2:sim_years){
  
  depletion <- purrr::map(mse_sim[[length(mse_sim)]], ~ sum(.x$ssb_p_a) / .x$ssb0)
  
  u <- map(depletion, ~  pmin(max_u,pmax(0.01,hcr_slope * .x + hcr_intercept)))
  
  quotas <- purrr::map2(mse_sim[[length(mse_sim)]], u, ~ sum(.x$ssb_p_a) * .y)
  
  next_year <- simmar(
  fauna = fauna,
  fleets = fleets,
  years = 1,
  initial_conditions = initial_conditions,
  manager = list(quotas = quotas),
  starting_step = starting_step,
  keep_starting_step = FALSE
)

  initial_conditions <- next_year[[length(next_year)]]
  
  starting_step <- as.numeric(last(names(next_year)))

  mse_sim <- append(mse_sim, next_year)
  
}

processed_mse_sim <- process_marlin(mse_sim)

plot_marlin(processed_mse_sim, plot_var = "c", max_scale = FALSE)

plot_marlin(processed_mse_sim, plot_var = "ssb", max_scale = FALSE)


```
