---
title: "optimize-mpas"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{optimize-mpas}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


One of the hardest questions of marine protected area (MPA) design is the simple "how big should it be and where should it go"? `marlin` can't tell you how big it should be, but it *can* help you evaluate the impacts to different objectives of different MPA networks, and if you supply it with an objective function weighting scheme, the best network to maximize your objective function. 

This is achieved with the `optimize_mpa` function, that at its core requires an `alpha` parameter that specifies the relative weighting of conservation to economic interests, where 1 means that the only objective is conservation, zero means the only objective is economic. 

`optimize_mpa` then finds the MPA network of a specified size that maximizes your objective function. 

```{r setup}
library(marlin)

library(tidyverse)

resolution <- 20

years <- 20

seasons <- 4

steps <- years * seasons

time_step <- 1 / seasons
# for now make up some habitat


skipjack_habitat <- expand_grid(x = 1:resolution, y = 1:resolution) %>%
  dplyr::mutate(habitat =  dnorm((x ^ 2 + y ^ 2), 20, 200)) %>% 
  pivot_wider(names_from = y, values_from = habitat) %>% 
  select(-x) %>% 
  as.matrix()


bigeye_habitat <- expand_grid(x = 1:resolution, y = 1:resolution) %>%
  mutate(habitat =  dnorm((x ^ 2 + y ^ 2), 300, 100)) %>% 
  pivot_wider(names_from = y, values_from = habitat) %>% 
  select(-x) %>% 
  as.matrix()


bigeye_habitat2 <- expand_grid(x = 1:resolution, y = 1:resolution) %>%
  mutate(habitat =  dnorm((x ^ .2 + y ^ .2), 100, 100)) %>% 
  pivot_wider(names_from = y, values_from = habitat) %>% 
  select(-x) %>% 
  as.matrix()

fauna <- 
  list(
    "skipjack" = create_critter(
      scientific_name = "Katsuwonus pelamis",
      base_habitat = list(skipjack_habitat, skipjack_habitat), # pass habitat as lists
      season_blocks = list(c(1, 2), c(3, 4)), # seasons each habitat apply to
      recruit_habitat = skipjack_habitat,
      adult_diffusion = 2, # standard deviation of the number of patches moved by adults
      seasons = seasons,
      init_explt = 0.4, 
      explt_type = "f"
      ),
    "bigeye" = create_critter(
      scientific_name = "Thunnus obesus",
      base_habitat = list(bigeye_habitat, bigeye_habitat2), # pass habitat as lists
      season_blocks = list(c(1, 2), c(3, 4)), # seasons each habitat apply to
      recruit_habitat = bigeye_habitat,
      adult_diffusion = 1,
      seasons = seasons,
      init_explt = 0.7, 
      explt_type = "f"
    )
  )

fleets <- list(
  "longline" = create_fleet(
    list(
       "skipjack" = Metier$new(
        critter = fauna$skipjack,
        price = 100,
        # price per unit weight
        sel_form = "logistic",
        # selectivity form, one of logistic or dome
        sel_start = .3,
        # percentage of length at maturity that selectivity starts
        sel_delta = .1,
        # additional percentage of sel_start where selectivity asymptotes
        catchability = .01,
        # overwritten by tune_fleet but can be set manually here
        p_explt = 1
      ),
       "bigeye" = Metier$new(
        critter = fauna$bigeye,
        price = 10,
        sel_form = "logistic",
        sel_start = .1,
        sel_delta = .01,
        catchability = 0,
        p_explt = 1
      )
    ),
    base_effort = resolution ^ 2,
        resolution = resolution

  ),
  "purseseine" = create_fleet(list(
    skipjack = Metier$new(
      critter = fauna$skipjack,
      price = 100,
      sel_form = "logistic",
      sel_start = 0.25,
      sel_delta = .1,
      catchability = .01,
      p_explt = 0.9
    ),
    bigeye = Metier$new(
      critter = fauna$bigeye,
      price = 100,
      sel_form = "logistic",
      sel_start = .25,
      sel_delta = .5,
      catchability = .01,
      p_explt = 1
    )
  ),
  base_effort = resolution ^ 2,    resolution = resolution
)
)

a <- Sys.time()

fleets <- tune_fleets(fauna, fleets) 

Sys.time() - a


# run simulations

# run the simulation using marlin::simmar
a <- Sys.time()

sim <- simmar(fauna = fauna,
                  fleets = fleets,
                  years = years)

Sys.time() - a
# a <- Sys.time()

processed_marlin <- process_marlin(sim = sim, time_step = time_step, keep_age = TRUE)
# Sys.time() - a

plot_marlin(processed_marlin)

plot_marlin(processed_marlin, plot_var = "c")

plot_marlin(processed_marlin, plot_var = "n", plot_type = "length", fauna = fauna)

plot_marlin(processed_marlin, plot_var = "ssb", plot_type = "space")

```


```{r}

mpa_network <-
  optimize_mpa(
    fauna = fauna,
    fleets = fleets,
    resolution = resolution,
    starting_conditions = sim[[length(sim)]],
    alpha = 0.75,
    max_prop_mpa = 0.1, 
    prop_sampled = 0.05,
    max_prop_mpa = 0.3,
    work_backwards = TRUE
  )


mpa_network$mpa_network %>% 
  group_by(p_protected) %>% 
  summarise(mean(mpa))

mpa_network$mpa_network %>% 
  # filter(p_protected  == 100) %>%
  ggplot(aes(x,y, fill = mpa)) + 
  geom_tile() + 
  facet_wrap(~p_protected)

```




