---
title: "Manage Catch and Effort of Fleets"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{fleet-management}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE
)
```


`marlin` allows for a wide range of options to govern both the management and internal dynamics of fishing fleets.

Things you can adjust include:

- **Fleet models**: `fleet_model` controls effort dynamics. Options are:
  - `"constant_effort"`: total effort fixed at `base_effort` each time step
  - `"open_access"`: effort adjusts based on average profitability, equilibrating where total profits = 0
  - `"sole_owner"`: effort adjusts based on marginal profitability, equilibrating at maximum economic yield (MEY) where marginal profit = 0
  - `"manual"`: effort taken from a user-supplied vector

- **Closed fishing seasons** per species (e.g. enforcing a closed season for species X but not species Y)

- **Catch quotas** per species

- **Effort caps** per fleet

- **Size limits and selectivity** forms per metier and species

- **No-take marine protected areas** (MPAs)

You can mix and match most of these options — for example, running an open-access fleet subject to a total quota for some species but not others.

First, let's set up the system and the species. In this case we use a simple example with one bigeye tuna population. 

```{r setup}
library(marlin)

library(tidyverse)

theme_set(theme_marlin(base_size = 14) + theme(legend.position = "top"))

resolution <- 10 # resolution is in squared patches, so 10 implies a 10x10 system, i.e. 100 patches

years <- 50

seasons <- 4

time_step <- 1 / seasons

steps <- years * seasons

fauna <- 
  list(
    "bigeye" = create_critter(
      scientific_name = "Thunnus obesus",
      adult_diffusion = 10,
      density_dependence = "post_dispersal", 
      seasons = seasons,
      depletion = 0.8,
      resolution = resolution,
      steepness = 0.6,
      ssb0 = 1000
    )
  )

fauna$bigeye$m_at_age
```


## Open access

Let's set up two fleets: one open access, one constant effort. Open-access dynamics are based on the profitability of the fishery and require a few more parameters, though reasonable defaults are provided.

### Open-access effort dynamics

Total effort adjusts each time step in response to a normalised profitability signal:

$$E_{t+1,f} = E_{t,f} \times \exp\left(\rho \cdot \tanh\left(\frac{s_t}{k}\right)\right)$$

where the profitability signal $s_t$ is:

$$s_t = \frac{R_t - C_t}{|R_t| + C_t}$$

and $R_t$ and $C_t$ are total revenue and total cost in the previous step, respectively. This signal is bounded in $[-1, 1]$, handles negative revenue safely, and is invariant to rescaling.

The parameters control entry/exit dynamics:

- $\rho$ (`oa_rho_year`): annual effort adjustment rate. A value of 0.5 means effort can grow or shrink by roughly 50% per year under strong profit or loss signals.
- $k$ (`oa_signal_half`): signal value at which the tanh response reaches half its maximum. Lower values make fleets more responsive. Typical range 0.15–0.4; default 0.3.
- `oa_max_growth_per_year`: maximum fractional increase in effort per year under highly profitable conditions (e.g. 0.5 = +50% per year).
- `oa_max_decline_per_year`: maximum fractional decrease in effort per year under highly unprofitable conditions (e.g. 0.5 = -50% per year).

The tanh function ensures that:

- Entry/exit speed saturates under extreme profits or losses (avoiding unrealistic boom/bust cycles).
- Response is approximately linear near break-even profitability.
- Effort converges to an equilibrium where total profits equal zero (the classic open-access outcome).

### Revenue and costs

Revenue is:

$$R_{t,f} = \sum_{s=1}^S p_{f,s} \, \text{Catch}_{f,s}$$

where $p$ is price and Catch is catch for species $s$ by fleet $f$.

Costs are:

$$C_{t,f} = c_0 \, E^{\text{ref}} \sum_{l=1}^P \left[\left(\frac{E_{t,l,f}}{E^{\text{ref}}}\right)^\gamma + \theta \, \tilde{d}_l \, \frac{E_{t,l,f}}{E^{\text{ref}}}\right]$$

where:

- $c_0$ (`cost_per_unit_effort`): base cost per unit of effort
- $E^{\text{ref}}$: reference effort per patch (typically `base_effort / n_patches`)
- $\gamma$ (`effort_cost_exponent`): exponent controlling congestion/convexity. Values > 1 make additional units of effort progressively more expensive. Default 1.2.
- $\theta$ (`travel_weight`): derived internally from `travel_fraction`
- $\tilde{d}_l$: normalised distance from patch $l$ to the nearest port

### Calibrating costs with `cr_ratio`

Rather than setting $c_0$ manually, specify a target cost-to-revenue ratio (`cr_ratio`) at equilibrium. A value of 1 means zero profits at equilibrium (classic open access), > 1 means losses, < 1 means positive profits.

`tune_fleets(..., tune_costs = TRUE)` then solves for the $c_0$ that achieves the desired `cr_ratio` at the tuned depletion level. 



```{r}

fleets <- list(
  "longline" = create_fleet(
    list("bigeye" = Metier$new(
        critter = fauna$bigeye,
        price = 10,
        sel_form = "logistic",
        sel_start = 1,
        sel_delta = .01,
        catchability = 0,
        p_explt = 2
      )
    ),
    base_effort = resolution ^ 2,
    resolution = resolution,
    cr_ratio = 1,
    travel_fraction = 0.5,
    fleet_model = "open_access",
    oa_max_growth_per_year = 0.5,
    oa_max_decline_per_year = 0.5,
    oa_signal_half = 0.3)
,
"handline" = create_fleet(
  list("bigeye" = Metier$new(
    critter = fauna$bigeye,
    price = 10,
    sel_form = "logistic",
    sel_start = 1,
    sel_delta = .01,
    catchability = 0,
    p_explt = 1
  )
  ),
  base_effort = resolution ^ 2,
  resolution = resolution,
  fleet_model = "constant_effort",
  cost_per_unit_effort = 2
))

fleets <- tune_fleets(fauna, fleets, tune_type = "depletion", tune_costs = TRUE) 
```


We can now run our simulation and examine the resulting fleet dynamics

```{r}

sim <- simmar(fauna = fauna,
                  fleets = fleets,
                  years = years)
proc_sim <- process_marlin(sim)

plot_marlin(proc_sim)

proc_sim$fleets %>% 
  group_by(step, fleet) %>%
  summarise(effort = sum(effort)) %>% 
  ggplot(aes(step * time_step, effort, color = fleet)) + 
  geom_line() + 
    scale_x_continuous(name = "Year")



```


## Open access and MPAs

The fleet model choice determines how effort responds to an MPA. Under the default **constant effort with reallocation** dynamics, total effort stays fixed but is redistributed from inside the MPA to the remaining fishable patches. Under the **open access** model, effort responds dynamically to how the MPA affects profitability.

When an MPA is implemented, the open-access fleet initially loses access to productive fishing grounds inside the closure. This reduces catch while costs remain the same, driving down profits and triggering effort contraction. As the MPA matures, spillover effects (adult movement and larval export from the protected area) can increase catch rates in adjacent patches, improving profitability and allowing effort to partially recover. Eventually the fleet settles at a new zero-profit equilibrium where the benefits of spillover are balanced against the lost access inside the MPA. 

```{r}

set.seed(42)
#specify some MPA locations
mpa_locations <- expand_grid(x = 1:resolution, y = 1:resolution) %>%
mutate(mpa = x > 4 & y < 6)

with_mpa <- simmar(fauna = fauna,
                  fleets = fleets,
                  years = years,
                  manager = list(mpas = list(locations = mpa_locations,
              mpa_year = floor(years * .5))))

proc_mpa_sim <- process_marlin(with_mpa)


proc_mpa_sim$fleets %>% 
  group_by(step, fleet) %>%
  summarise(effort = sum(effort)) %>% 
  ggplot(aes(step * time_step, effort, color = fleet)) + 
  geom_line() + 
  scale_x_continuous(name = "year")


```


## Sole owner

The `"sole_owner"` fleet model is identical to open access in its effort adjustment mechanics, but uses the *marginal* profit signal rather than the *average* profit signal. This causes the fleet to equilibrate at maximum economic yield (MEY), where marginal profit equals zero, rather than at the open-access equilibrium where total profit equals zero.

Marginal profit is computed via finite-difference perturbations (see `?calc_marginal_value`). This requires additional computation each step, so `"sole_owner"` fleets run slightly slower than `"open_access"` fleets. The equilibrium outcome typically features lower effort and higher biomass than open access, since the fleet stops expanding when the *next* unit of effort would no longer be profitable, rather than waiting until *all* effort is unprofitable.

To use sole owner dynamics:

```{r, eval = FALSE}
fleets <- list(
  "longline" = create_fleet(
    list("bigeye" = Metier$new(...)),
    fleet_model = "sole_owner",
    cr_ratio = 0.8,  # MEY typically has positive profits, so cr_ratio < 1
    ...
  )
)
```

## Quotas

Quotas can be layered onto any fleet model. Here we impose a total catch quota of 8 units of bigeye across all fleets. The quota is enforced each time step by solving for a scalar effort multiplier that keeps total catch at or below the quota.

**Important:** quotas impose a *cap*, not a requirement. In the early days of the fishery when catches would naturally exceed the quota, the quota is binding and effort is reduced proportionally. However, in later years when the population has declined and effort dynamics (especially under open access) would produce catches below the quota anyway, the fleet is unconstrained and the quota has no effect. This reflects real-world quota fisheries where economic or biological conditions can make quotas non-binding. 

```{r}
sim_quota <- simmar(fauna = fauna,
                  fleets = fleets,
                  years = years,
                  manager = list(quotas = list(bigeye = 8)))

proc_sim_quota <- process_marlin(sim_quota)

plot_marlin(proc_sim_quota, plot_var = "c", max_scale = FALSE)

proc_sim_quota$fleets %>% 
  group_by(step, fleet) %>% 
  summarise(catch = sum(catch)) %>% 
  ggplot(aes(step * time_step, catch, color = fleet)) + 
  geom_line()+
    scale_x_continuous(name = "Year")


proc_sim_quota$fleets %>% 
  group_by(step, fleet) %>%
  summarise(effort = sum(effort)) %>% 
  ggplot(aes(step * time_step, effort, color = fleet)) + 
  geom_line() + 
    scale_x_continuous(name = "Year")


```

## Effort caps

Another management option is to set a maximum total effort per fleet. This could reflect regulation (e.g. limited entry permits) or physical reality (e.g. if effort is measured in "days fished per year" by a fixed number of vessels, there is a natural ceiling).

Set effort caps via `manager = list(effort_cap = list(FLEET_NAME = EFFORT_CAP))`, where `FLEET_NAME` is the name of the fleet and `EFFORT_CAP` is the maximum total effort allowed.

**Note:** effort caps are only meaningful when `fleet_model == "open_access"` or `"sole_owner"`. For constant-effort fleets, effort is already fixed at `base_effort`. Under open-access or sole-owner dynamics, the cap ensures that while profitability signals can *reduce* total effort, effort can never expand beyond the cap. The fleet can still contract below the cap if losses warrant it, but profitable conditions will not drive effort above the ceiling. 

```{r}

cap = 1.1*fleets$longline$base_effort

sim_effort <- simmar(fauna = fauna,
                  fleets = fleets,
                  years = years,
                  manager = list(effort_cap = list(longline = cap)))

proc_sim_effort <- process_marlin(sim_effort)

plot_marlin(proc_sim_effort, plot_var = "c", max_scale = FALSE)

proc_sim_effort$fleets %>% 
  group_by(step, fleet) %>% 
  summarise(catch = sum(catch)) %>% 
  ggplot(aes(step * time_step, catch, color = fleet)) + 
  geom_line()+
    scale_x_continuous(name = "Year")


proc_sim_effort$fleets %>% 
  group_by(step, fleet, patch) %>%
  summarise(effort = unique(effort)) %>% 
  group_by(step, fleet) |> 
  summarise(effort = sum(effort)) |> 
  ggplot(aes(step * time_step, effort, color = fleet)) + 
  geom_line() + 
  geom_hline(yintercept = cap) +
    scale_x_continuous(name = "Year") + 
  scale_y_continuous(limits = c(0, NA))


```



<!-- ## Seasons -->

<!-- We can also impose closed or reduced fishing seasons that apply to one or more targeted species. This is specified by a vector containing values between 0 and 1 with length equal to the number of seasons in the simulation.  When a season has a value of 1, that implies that fishing continues as normal in that season. When that value is 0, that means that the fishery is closed in that season. Setting intermediate values, like say 0.5, is akin to simulating a closed seasons with a 50% discard mortality rate -->

<!-- So one option would be to specify a discard mortality rate, and a discard rate, with the discard rate being potentially seasonal.  -->

<!-- The problem is the need to separate out the accounting between how many fish actually die and how many the fleet is allowed to retain. Suppose you set a discard rate of 50%, and a discard mortality rate of 0%. That is equivalent to just cutting effort in half.  -->

<!-- But, suppose you set a discard rate of 100% and a discard mortality rate of 100%. That would mean that the same amount of fish are dying, but you can't keep them. So that would be easy since again effort doesn't change, you'd just need a multiplier on converting dead fish to catch.  -->

<!-- But, now suppose that discard rate is 100% and discard mortality is 50%. Now, you need to both reduce effort by 50%, but allow zero retention.  -->

<!-- So, you'd need to adjust fishing mortality rate by a function of discard rate + discard mortality -->

<!-- That will get the correct amount of fish dead -->

<!-- Then you'd adjust the catch by just the discard rate -->

<!-- This gets pretty complicated pretty quickly though, since it means that you need to adjust the actual amount killed. So that would come in on a multiplier of the fishing mortality rate in the manner you have now, where say if you  -->


<!-- ```{r} -->

<!-- sim <- simmar(fauna = fauna, -->
<!--                   fleets = fleets, -->
<!--                   years = years, -->
<!--               manager = list(seasonal_effort_mult = list(bigeye = c(1,2)))) -->

<!-- proc_sim <- process_marlin(sim) -->

<!-- plot_marlin(proc_sim, plot_var = "c") -->

<!-- proc_sim$fleets %>%  -->
<!--   group_by(step, fleet) %>% -->
<!--   summarise(effort = sum(effort)) %>%  -->
<!--   ggplot(aes(step * time_step, effort, color = fleet)) +  -->
<!--   geom_line() +  -->
<!--   scale_x_continuous(name = "Year") +  -->
<!--   scale_y_continuous(limits = c(0, NA)) -->


<!-- ``` -->





## Manual effort

As an alternative to model-driven effort dynamics, you can manually specify total effort for each time step. Set `fleet_model = "manual"` and provide a vector of effort values in the fleet object's `effort` field. The vector must have length equal to `years * seasons`.

This is useful when you want to impose an exogenous effort trajectory — for example, to simulate the historical expansion of a fishery, test the effects of a known effort time series, or explore counterfactual scenarios where effort follows a prescribed pattern rather than responding endogenously to economic signals. 

```{r}
time_steps <- years * seasons

fleets$longline$fleet_model <- "manual"

fleets$longline$effort <- fleets$longline$base_effort *  rlnorm(time_steps,0,.2)
  
fleets <- tune_fleets(fauna, fleets, tune_type = "depletion")

sim_effort <- simmar(
  fauna = fauna,
  fleets = fleets,
  years = years
)

proc_sim_effort <- process_marlin(sim_effort)

plot_marlin(proc_sim_effort, plot_var = "c", max_scale = FALSE)

proc_sim_effort$fleets %>%
  group_by(step, fleet) %>%
  summarise(catch = sum(catch)) %>%
  ggplot(aes(step * time_step, catch, color = fleet)) +
  geom_line() +
  scale_x_continuous(name = "Year")


proc_sim_effort$fleets %>%
  group_by(step, fleet) %>%
  summarise(effort = sum(effort)) %>%
  ggplot(aes(step * time_step, effort, color = fleet)) +
  geom_line() +
  scale_x_continuous(name = "Year")

```




<!-- ## Seasons -->


<!-- We can also impose closed or reduced fishing seasons that apply to one or more targeted species. This is specified by a vector containing values between 0 and 1 with length equal to the number of seasons in the simulation.  When a season has a value of 1, that implies that fishing continues as normal in that season. When that value is 0, that means that the fishery is closed in that season. Setting intermediate values, like say 0.5, is akin to simulating a closed seasons with a 50% discard mortality rate -->

<!-- So one option would be to specify a discard mortality rate, and a discard rate, with the discard rate being potentially seasonal.  -->

<!-- The problem is the need to separate out the accounting between how many fish actually die and how many the fleet is allowed to retain. Suppose you set a discard rate of 50%, and a discard mortality rate of 0%. That is equivalent to just cutting effort in half.  -->

<!-- But, suppose you set a discard rate of 100% and a discard mortality rate of 100%. That would mean that the same amount of fish are dying, but you can't keep them. So that would be easy since again effort doesn't change, you'd just need a multiplier on converting dead fish to catch.  -->

<!-- But, now suppose that discard rate is 100% and discard mortality is 50%. Now, you need to both reduce effort by 50%, but allow zero retention.  -->

<!-- So, you'd need to adjust fishing mortality rate by a function of discard rate + discard mortality -->

<!-- That will get the correct amount of fish dead -->

<!-- Then you'd adjust the catch by just the discard rate -->

<!-- This gets pretty complicated pretty quickly though, since it means that you need to adjust the actual amount killed. So that would come in on a multiplier of the fishing mortality rate in the manner you have now, where say if you  -->


<!-- ```{r} -->

<!-- sim <- simmar(fauna = fauna, -->
<!--                   fleets = fleets, -->
<!--                   years = years, -->
<!--               manager = list(seasonal_effort_mult = list(bigeye = c(1,2)))) -->

<!-- proc_sim <- process_marlin(sim) -->

<!-- plot_marlin(proc_sim, plot_var = "c") -->

<!-- proc_sim$fleets %>%  -->
<!--   group_by(step, fleet) %>% -->
<!--   summarise(effort = sum(effort)) %>%  -->
<!--   ggplot(aes(step * time_step, effort, color = fleet)) +  -->
<!--   geom_line() +  -->
<!--   scale_x_continuous(name = "Year") +  -->
<!--   scale_y_continuous(limits = c(0, NA)) -->


<!-- ``` -->



