---
title: "Invertebrates"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(marlin)

library(tidyverse)

library(tictoc)

library(ggridges)

theme_set(theme_marlin())
```

Invertebrates have some slightly different life history than finfish. `marlin` has some capabilities to simulate some of these dynamics, but be warned that the less like a finfish, the less appropriate `marlin` may be. 

First, let's set up parameters. Since invertebrates are not included in `FishLife`, all required life history values must be supplied manually. 

For invertebrates, we are going to implement a couple of features. 

- We are going to run the simulation on a weekly timestep, to better approximate a fast growing species

- We are going to use a power function to model size at age, to approximate a critter that simply expands in volume as it ages, rather than reaching an asymptotic size

- We will make the species semelparous, meaning that individuals die after spawning. This is implemented by introducing a mortality term in which `1 - (proportion_mature_at_age)` individuals in each age group die **after** spawning


```{r}
resolution <- 5 # the resolution of the system

patch_area <- 1 # the area of each patch

kp <- .5 # the patchiness of habitat

years <- 10 # number of years to run simualtion

seasons <- 52 # weekly time step

time_step <- 1 / seasons

max_age <- 1.5 # the maximum age of the animal (in years)

length_bin_width <- 0.1 # the width of each length bin in centimeters

length_a <- 0.1 # scale for growth power function

length_b <- 3 # exponent for growth power function. Consider a spherical octopus.

lorenzen_c <- -1 # the rate of the lorenzen natural mortality curve. More negative values increase natural mortality at age 0 relative to max age.

m <- 0.4 # asymptotic natural mortality

t0 <- -.5 # size at age 0

ages <- seq(0, max_age, by = time_step)

length_at_age <- length_a * (ages - t0)^length_b

max_length <- max(length_at_age)

m_at_age <- m * (length_at_age / max(length_at_age))^lorenzen_c

critter_correlations <- matrix(c(1, -.5, -.5, 1), nrow = 2, byrow = TRUE) # simulate spatial correlation between species

species_distributions <- sim_habitat(
  critters = c("squishy", "squashy"),
  resolution = resolution,
  patch_area = patch_area,
  kp = kp,
  output = "list",
  critter_correlations = critter_correlations
)
```


```{r}
species_distributions$critter_distributions |>
  map(as.data.frame) |>
  list_rbind(names_to = "critter") |>
  group_by(critter) |>
  mutate(x = 1:n()) |>
  ungroup() |>
  pivot_longer(-c(critter, x), names_to = "y", values_to = "habitat") |>
  group_by(critter) |>
  mutate(habitat = habitat / max(habitat)) |>
  ungroup() |>
  ggplot(aes(x, y, fill = habitat)) +
  geom_tile() +
  facet_wrap(~critter) +
  scale_fill_viridis_c()
```

Next, we'll set up the fauna and fleet objects

```{r}
fauna <-
  list(
    "squishy" = create_critter(
      spawning_seasons = c(1:20),
      habitat = species_distributions$critter_distributions$squishy,
      common_name = "squishy",
      growth_model = "power",
      length_a = length_a,
      length_b = length_b,
      length_bin_width = length_bin_width,
      t0 = t0,
      cv_len = 0.6,
      m_at_age = m_at_age,
      max_age = max_age,
      age_mature = max_age - .5,
      semelparous = TRUE,
      delta_mature = .1,
      weight_a = 2,
      weight_b = 1,
      adult_diffusion = .1,
      recruit_diffusion = 10,
      init_explt = 1.2,
      resolution = resolution,
      sigma_rec = 0,
      ac_rec = .3,
      query_fishlife = FALSE,
      seasons = seasons
    ),
    "squashy" = create_critter(
      spawning_seasons = c(21:40),
      habitat = species_distributions$critter_distributions$squashy,
      common_name = "squashy",
      growth_model = "power",
      length_a = length_a,
      length_b = 2,
      length_bin_width = length_bin_width,
      t0 = t0,
      cv_len = 0.6,
      m_at_age = m_at_age,
      max_age = max_age,
      age_mature = max_age / 2,
      semelparous = FALSE,
      delta_mature = .1,
      weight_a = 2,
      weight_b = 1,
      adult_diffusion = .1,
      recruit_diffusion = 10,
      init_explt = .6,
      resolution = resolution,
      sigma_rec = .6,
      ac_rec = .3,
      query_fishlife = FALSE,
      seasons = seasons
    )
  )

fleets <- list(
  "artisanal" = create_fleet(list(
    "squishy" = Metier$new(
      critter = fauna$squishy,
      p_explt = 0.5,
      sel_unit = "p_of_mat",
      sel_start = 0.1,
      sel_delta = 0.1
    ),
    "squashy" = Metier$new(
      critter = fauna$squashy,
      p_explt = 0.5,
      sel_unit = "p_of_mat",
      sel_start = 0.1,
      sel_delta = 0.1
    )
  ), resolution = resolution),
  "commercial" = create_fleet(list(
    "squishy" = Metier$new(
      critter = fauna$squishy,
      p_explt = 0.5,
      sel_unit = "length",
      sel_start = 0.25 * max_length,
      sel_delta = .01
    ),
    "squashy" = Metier$new(
      critter = fauna$squashy,
      p_explt = 0.5,
      sel_unit = "p_of_mat",
      sel_start = 0.5,
      sel_delta = 0.1
    )
  ), resolution = resolution)
)

fleets <- tune_fleets(fauna, fleets, tune_type = "explt", years = 20)
```

```{r}
fauna$squishy$plot()
```

```{r}
fauna$squashy$plot()
```


```{r}
sels <- data.frame(
  artisinal = fleets$artisanal$metiers$squishy$sel_at_length,
  commercial = fleets$commercial$metiers$squishy$sel_at_length,
  length = as.numeric(colnames(fauna$squishy$length_at_age_key))
)

sels |>
  pivot_longer(-length, names_to = "fleet", values_to = "selectivity") |>
  ggplot(aes(length, selectivity, color = fleet)) +
  geom_line()
```

```{r}
tic()
squishy_sim <- simmar(fauna, fleets, years = years, cor_rec = critter_correlations)
toc()
processed_squishy <- process_marlin(sim = squishy_sim, time_step = time_step)
```


```{r}
processed_squishy$fauna |>
  group_by(step, age, mean_length) |>
  summarise(number = sum(n)) |>
  group_by(step) |>
  mutate(pn = number / sum(number)) |>
  ungroup() |>
  ggplot(aes(mean_length, factor(step), height = pn)) +
  geom_ridgeline(stat = "identity", color = "transparent", fill = "red", scale = 10) +
  theme(axis.text.y = element_blank(), axis.ticks.y = element_blank())
```


```{r fig-rec,fig.cap="Number of recruits over time per species."}
processed_squishy$fauna |>
  filter(age == min(age)) |>
  ggplot(aes(step, n, color = critter)) +
  geom_line()
```

```{r fig-b,fig.cap="Biomass per species over time."}
plot_marlin(processed_squishy, fauna = fauna, plot_var = "b", max_scale = FALSE)
```

```{r, fig.cap="Example age composition over the course of a year, along with maturity ogive. Note one species is semelparous, the other is not."}
b <- processed_squishy$fauna |>
  group_by(year, age, critter) |>
  summarise(n = sum(n)) |>
  group_by(critter, year) |>
  mutate(yn = n / max(n)) |>
  ungroup()

mogive <- map(fauna, \(x) tibble(age = x$ages, maturity = x$maturity_at_age, semelparous = x$semelparous)) |>
  list_rbind(names_to = "critter")

b |>
  filter(year == max(year)) |>
  ggplot() +
  geom_col(aes(age, yn, fill = "Age Composition")) +
  geom_line(data = mogive, aes(age, maturity, linetype = semelparous, color = "Maturity At Age")) +
  facet_wrap(~critter) +
  scale_fill_manual(name = "", values = "blue") +
  scale_color_discrete(name = "") +
  scale_y_continuous(name = "Annual Age Composition")
```

```{r, fig.cap="Age composition over time"}
a <- processed_squishy$fauna |>
  group_by(year, step, age, critter) |>
  summarise(n = sum(n)) |>
  ungroup() |>
  group_by(critter, year, step) |>
  mutate(sn = n / max(n)) |>
  group_by(critter, year) |>
  mutate(n = n / max(n)) |>
  ungroup()

a |>
  filter(year < 4) |>
  ggplot() +
  geom_density_ridges(aes(age, step, height = sn, group = step, fill = step),
    stat = "identity",
    color = "transparent",
    scale = 4,
    alpha = 0.5
  ) +
  facet_wrap(~critter, scales = "free_x") +
  scale_fill_viridis_c() +
  scale_y_continuous(name = "Time Step")
```

