---
title: "Spatial Allocation Strategies"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{spatial-allocation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 5
)
set.seed(42)
```

When a fleet decides *where* to fish, it optimizes some objective across the available patches. In `marlin`, the `spatial_allocation` argument to `create_fleet` controls what that objective is. Different objectives produce very different spatial footprints of fishing effort, which in turn shape the spatial depletion patterns for every species in the system.

`marlin` currently supports the following `spatial_allocation` options:

- **`rpue`** (revenue per unit effort): the fleet chases the highest catch rates weighted by price. This is the default and a common assumption in spatial fleet models.
- **`revenue`**: the fleet moves toward patches with the highest *total* revenue, not per-unit-effort. High-effort patches stay attractive even as catch rates decline.
- **`ppue`** (profit per unit effort): like `rpue`, but subtracts the cost of fishing (including travel costs from port). Patches far from port become less attractive.
- **`profit`**: the fleet moves toward patches with the highest *total* profit. Similar to `revenue` but cost-aware.
- **`marginal_profit`**: the fleet responds to the *marginal* return of adding one more unit of effort to a patch. This is the most economically sophisticated option: a patch with high total profit but diminishing returns will be less attractive than one where the next unit of effort still pays off.
- **`cpue`** / **`catch`**: catch-based variants (not price-weighted).
- **`marginal_revenue`**: like `marginal_profit` but ignoring costs.
- **`manual`**: effort distributed proportionally to continuous weights in `fishing_grounds$fishing_ground`.
- **`uniform`**: effort spread equally across all open patches.

The mechanics of how effort is redistributed in response to these objectives is explained in detail in the "How spatial allocation works" section below. This vignette then demonstrates each option using a shared two-species, two-fleet system with fishing ports, so that the spatial cost structure matters for the cost-aware strategies.

Independently of the strategy, fleets can optionally smooth their objective surface over time via the `objective_memory_halflife` argument to `create_fleet()`. This dampens single-step feedback oscillations under strong fleet-biomass coupling (e.g. after MPA closures). It is documented in "Optional: temporal memory of the objective surface" below, with an empirical demonstration in Scenario 5.

## How spatial allocation works: the algorithm

Spatial effort allocation in `marlin` is handled by the `allocate_effort` function, which uses a multiplicative velocity-field update to redistribute a fleet's total effort across patches in response to an economic signal (the "objective"). The algorithm is designed to be robust, gradient-based, and numerically stable even when patches differ wildly in profitability.

### Step-by-step procedure

For each fleet at each time step:

**1. Flatness detection**  
Before attempting to optimize, the algorithm checks whether the objective is effectively uniform across patches. If the range-based coefficient of variation is below a threshold (default 0.1%):

$$\text{CV}_{\text{range}} = \frac{\max(O) - \min(O)}{|\text{median}(O)|} < \text{flatness\_tol}$$

then effort is distributed uniformly across all open patches and the optimization is skipped. This prevents spurious concentration along edges when there is no meaningful gradient to follow.

**2. Standardization**  
The objective $O_p$ in each patch $p$ is centered and scaled to produce a $z$-score:

$$z_p = \frac{O_p - \text{median}(O)}{\text{scale}(O)}$$

where $\text{scale}(O)$ can be the median absolute deviation (default, most robust), interquartile range, or standard deviation. The $z$ scores are then clipped to $[-c, c]$ (default $c = 5$) to prevent extreme updates from outliers:

$$z_p \leftarrow \max(-c, \min(c, z_p))$$

Closed patches have $z_p = 0$.

**3. Velocity field**  
The standardized objective becomes a velocity after removing the mean across open patches:

$$v_p = z_p - \bar{z}_{\text{open}}$$

This ensures that the velocity field integrates to zero over the open domain, which is necessary for conserving total effort. Patches with above-average objectives have $v_p > 0$ (attracting effort), and patches with below-average objectives have $v_p < 0$ (repelling effort).

**4. Multiplicative update**  
Current effort $e_p$ is updated via:

$$e_p^{\text{new}} \propto e_p \exp(\eta \, v_p)$$

where $\eta$ (default 1.0) controls the step size. The proportionality constant is chosen to conserve total effort:

$$e_p^{\text{new}} = e_p \exp(\eta \, v_p) \times \frac{E_{\text{total}}}{\sum_p e_p \exp(\eta \, v_p)}$$

This is implemented in log-space for numerical stability. Closed patches receive $e_p^{\text{new}} = 0$.

**5. Optional exploration mixing**  
If `eps_mix` > 0, the result is blended with a uniform distribution to allow re-entry into temporarily abandoned patches:

$$e_p^{\text{final}} = (1 - \epsilon) \, e_p^{\text{new}} + \epsilon \, \frac{E_{\text{total}}}{N_{\text{open}}}$$

### Optional: temporal memory of the objective surface

`objective_memory_halflife` is a **conditional remedy, not a default upgrade**. If you see period-2 sawtooth in patch-level effort or catch (typically under `marginal_profit` / `marginal_revenue` allocation, or with MPA closures + cost-aware allocation), it can dampen the oscillation. If your run looks smooth at the default `halflife = 0`, leave it there.

The mechanism: by default, the objective surface a fleet acts on is rebuilt fresh each step from the previous step's outcomes — a one-step-lag signal. Under strong fleet-biomass coupling this can drive period-2 sawtooth oscillation in patch effort: last-step buffet picks where to fish, this-step fishing changes biomass, next-step buffet inverts. The vanilla scenarios in this vignette don't trigger this, but structural perturbations like MPA closures (which suddenly shift the open-patch set) can.

Setting `objective_memory_halflife > 0` on a fleet replaces the raw per-patch objective $O_p$ with an exponentially-weighted moving average:

$$\tilde{O}_p^{(t)} = \alpha \, O_p^{(t)} + (1 - \alpha) \, \tilde{O}_p^{(t-1)}, \qquad \alpha = 1 - 0.5^{1 / h_{\text{steps}}}$$

The half-life is specified in **years** and converted internally to steps: $h_{\text{steps}} = h_{\text{years}} \times \text{steps\_per\_year}$, so a given value produces the same calendar-time smoothing regardless of how many seasons per year the model uses.

Three implementation details matter:

- **Closed-patch policy** is "freeze and resume": the EWMA only updates on patches that are currently open. Closed patches retain their last-seen smoothed value, so an MPA that closes then re-opens a patch sees the fleet's pre-closure memory of it.
- **Bootstrap ramp**: early post-bootstrap steps use $\alpha_{\text{eff}} = \max(\alpha, 1/n)$ where $n$ counts post-bootstrap updates for that fleet. This avoids anchoring the smoothed surface to the first observed buffet column.
- **Phase-lag tradeoff**: an EWMA introduces a phase lag of roughly $1.44 \times h_{\text{years}}$ years between a real change in patch marginal value and the fleet's perceived value. Long half-lives can replace high-frequency sawtooth with low-frequency overshoot — a different pathology. Practical sweet spot is **0.5–1.5 years** for most fishery time scales; see Scenario 5 for an empirical demonstration.

Memory smooths only the *spatial allocation* objective. Total fleet effort under `open_access` or `sole_owner` still responds to the raw fleet-level profitability signal, which has its own slow timescale built in via the entry/exit caps.

### Why this algorithm?

The multiplicative update has several desirable properties:

- **Gradient-based**: Effort flows "uphill" toward high-value patches along the objective gradient.
- **Effort-conserving**: Total effort $\sum_p e_p$ is exactly preserved.
- **Non-negative**: Effort cannot become negative (provided it starts non-negative).
- **Smooth transitions**: The exponential update avoids discontinuous jumps, making the dynamics more realistic.
- **Scale-invariant**: The standardization step ensures that the update magnitude is comparable across fleets and time steps regardless of absolute objective values.

The algorithm is related to replicator dynamics from evolutionary game theory and gradient flow on the simplex. In the limit of small $\eta$, the continuous-time equivalent is:

$$\frac{de_p}{dt} = e_p \left( v_p - \langle v \rangle \right)$$

where $\langle v \rangle = \sum_p (e_p / E_{\text{total}}) v_p$ is the effort-weighted average velocity. This is a stable dynamical system that converges to a Nash equilibrium where effort is concentrated in patches with the highest objective values.

### Objectives available

The objective $O_p$ is drawn from the "buffet" returned by `go_fish`, which computes prospective catch, revenue, and profit for each patch without actually removing fish. The `spatial_allocation` setting maps to:

| `spatial_allocation` | Objective \(O_p\) | Description | Economic interpretation |
|---|---|---|---|
| `rpue` | Revenue per unit effort | Value-weighted catch rate | Effort shifts toward patches that produce higher revenue per unit effort. This reflects fleets choosing locations where each unit of effort generates more revenue on average, without explicitly accounting for how crowding reduces returns. |
| `revenue` | Total revenue | Value-weighted catch | Effort shifts toward patches that produce the highest total revenue. This reflects fleets targeting areas with the highest overall value of catch, regardless of operating costs or how returns change as more vessels enter the area. |
| `ppue` | Profit per unit effort | Revenue net of costs per effort | Effort shifts toward patches that produce higher profit per unit effort. This reflects fleets choosing locations where each unit of effort generates more net profit on average, accounting for travel and operating costs. |
| `profit` | Total profit | Revenue minus costs | Effort shifts toward patches that produce the highest total profit. This reflects fleets targeting areas with the highest overall net returns, without explicitly accounting for how additional effort changes profitability. |
| `cpue` | Catch per unit effort | Biomass-weighted catch rate | Effort shifts toward patches that produce higher catch per unit effort. This reflects fleets choosing locations where each unit of effort generates more catch on average, regardless of price. |
| `catch` | Total catch | Biomass-weighted catch | Effort shifts toward patches that produce the highest total catch. This reflects fleets targeting areas with the highest total biomass or catch, regardless of economic value. |
| `marginal_profit` | Marginal profit | Change in total profit from the next unit of effort (finite difference) | Effort shifts toward patches where adding a small amount of additional effort would increase total profit the most. Over time, effort tends to distribute so that the last unit of effort earns similar profit across all actively fished patches. This is the closest approximation to the **Ideal Free Distribution (IFD)**, where effort distributes so that no vessel can increase profit by moving effort to another patch. |
| `marginal_revenue` | Marginal revenue | Change in total revenue from the next unit (finite difference) | Effort shifts toward patches where adding a small amount of effort would increase total revenue the most. Over time, effort tends to distribute so that the last unit of effort generates similar revenue across actively fished patches. This reflects the production-driven component of the **Ideal Free Distribution (IFD)**, but does not fully account for costs. |
| `manual` | User weights | Proportional to `fishing_grounds$fishing_ground` | Effort follows externally specified spatial preferences (e.g., fishing traditions, regulations, or predefined fishing grounds), rather than responding directly to economic outcomes. |
| `uniform` | None | Equal effort across open patches | Effort is spread evenly across space. Useful as a baseline representing no spatial preference or complete uncertainty about conditions. |


**Choosing a strategy for your use case:**

- **Unregulated open-access fishery?** → `revenue` or `rpue` (strong tragedy-of-commons)
- **Limited-entry or regulated fishery with individual decisions?** → `ppue` (moderate, cost-aware)
- **Cooperative, ITQ system, or single-owner fleet?** → `marginal_profit` (sole-owner optimum)
- **Transitional or mixed governance?** → `profit` or `marginal_revenue`
- **Exploring spatial management or MPA impacts across behavioral regimes?** → Run multiple scenarios to bracket uncertainty

The marginal variants require `calc_marginal_value` to compute finite-difference derivatives each step, which adds computational cost but produces more economically sophisticated spatial behavior (especially for sole-owner fleets).

## Shared Setup

We'll use a 10×10 grid with two critters (bigeye tuna and skipjack tuna) that have different spatial habitats generated by `sim_habitat`, and two fishing fleets (a longline fleet under open access dynamics and a handline fleet under constant effort). Two ports are placed asymmetrically to create spatial cost gradients.

```{r setup}
library(marlin)
library(ggplot2)
library(dplyr)
library(tidyr)

theme_set(marlin::theme_marlin(base_size = 12) +
            theme(legend.position = "top"))

resolution <- c(20, 20)
patches <- prod(resolution)
years <- 20
seasons <- 2
time_step <- 1 / seasons

# Two ports: one nearshore corner, one offshore
ports <- data.frame(x = c(1, 8), y = c(1, 9))
```

### Habitat

We use `sim_habitat` to generate spatially correlated habitat for each species. The `kp` parameter controls the spatial smoothness (lower = smoother), and we supply a negative cross-species correlation so that the two species prefer different parts of the seascape. This creates a spatial tradeoff: fleets cannot maximize catch of both species in the same patch.

```{r habitat}
# Negative correlation: species prefer different areas
critter_correlations <- matrix(c(1, -0.5,
                                 -0.5, 1), nrow = 2)

habitats <- sim_habitat(
  critters = c("bigeye", "skipjack"),
  kp = 0.1,
  critter_correlations = critter_correlations,
  resolution = resolution,
  patch_area = 1,
  output = "list"
)

# Extract the habitat matrices
bigeye_habitat <- habitats$critter_distributions$bigeye
skipjack_habitat <- habitats$critter_distributions$skipjack

# Quick look at the generated habitats
habitat_df <- expand_grid(critter = c("bigeye", "skipjack"),
                          x = 1:resolution[1],
                          y = 1:resolution[2]) %>%
  arrange(critter, x, y) %>%
  mutate(
    habitat = c(as.vector(bigeye_habitat),
                as.vector(skipjack_habitat))
  )

ggplot(habitat_df) +
  geom_tile(aes(x, y, fill = habitat)) +
  geom_point(data = ports, aes(x, y),
             color = "red", size = 3, shape = 17) +
  scale_fill_viridis_c(option = "D") +
  facet_wrap(~critter) +
  coord_equal() +
  labs(title = "Underlying Habitat Quality",
       subtitle = "Red triangles = port locations",
       fill = "Habitat")
```

### Fauna

Both species use the habitats generated above. Bigeye are a slower-moving, more heavily fished species, while skipjack are more mobile and less depleted.

```{r fauna}
fauna <-
  list(
    "bigeye" = create_critter(
      common_name = "bigeye tuna",
      habitat = list(bigeye_habitat),
      season_blocks = list(1:seasons),
      adult_home_range = 5,
      recruit_home_range = 10,
      density_dependence = "local_habitat",
      seasons = seasons,
      depletion = 0.5,
      init_explt = 0.2,
      explt_type = "f",
      resolution = resolution,
      steepness = 0.6,
      ssb0 = 1000
    ),
    "skipjack" = create_critter(
      scientific_name = "Katsuwonus pelamis",
      habitat = list(skipjack_habitat),
      season_blocks = list(1:seasons),
      adult_home_range = 3,
      recruit_home_range = 8,
      density_dependence = "local_habitat",
      seasons = seasons,
      depletion = 0.6,
      init_explt = 0.15,
      explt_type = "f",
      resolution = resolution,
      steepness = 0.7,
      ssb0 = 800
    )
  )
```

### Fleet Builder

To keep the code DRY, we define a helper function that creates the two-fleet system for any pair of `spatial_allocation` settings. The longline fleet operates under open access (total effort responds to profitability) and the handline fleet under constant effort. Both fleets have the same port locations and a `travel_fraction` of 0.3, meaning travel costs comprise 30% of total costs at equilibrium.

```{r fleet-builder}
build_fleets <- function(longline_allocation, handline_allocation) {

  fleets <- list(
    "longline" = create_fleet(
      list(
        "bigeye" = Metier$new(
          critter = fauna$bigeye,
          price = 10,
          sel_form = "logistic",
          sel_start = 1,
          sel_delta = 0.01,
          catchability = 0,
          p_explt = 2
        ),
        "skipjack" = Metier$new(
          critter = fauna$skipjack,
          price = 5,
          sel_form = "logistic",
          sel_start = 0.8,
          sel_delta = 0.1,
          catchability = 0,
          p_explt = 1
        )
      ),
      ports = ports,
      base_effort = patches,
      resolution = resolution,
      spatial_allocation = longline_allocation,
      fleet_model = "open_access",
      cr_ratio = 1,
      travel_fraction = 0.7
    ),
    "handline" = create_fleet(
      list(
        "bigeye" = Metier$new(
          critter = fauna$bigeye,
          price = 10,
          sel_form = "logistic",
          sel_start = 1.2,
          sel_delta = 0.1,
          catchability = 0,
          p_explt = 1
        ),
        "skipjack" = Metier$new(
          critter = fauna$skipjack,
          price = 8,
          sel_form = "logistic",
          sel_start = 0.5,
          sel_delta = 0.2,
          catchability = 0,
          p_explt = 2
        )
      ),
      ports = ports,
      base_effort = patches,
      resolution = resolution,
      spatial_allocation = handline_allocation,
      fleet_model = "constant_effort",
      cr_ratio = 1,
      travel_fraction = 0.5
    )
  )

  fleets <- tune_fleets(fauna, fleets, tune_type = "depletion")

  fleets
}
```

### Plotting Helpers

We define helpers to extract the final-step effort and biomass and plot them as spatial maps with port locations.

```{r plot-helpers}
plot_effort <- function(sim, fleet_name, title = "") {
  final_step <- sim[[length(sim)]]
  effort_vec <- final_step[[1]]$e_p_fl[, fleet_name]

  patch_effort <- expand_grid(
    x = 1:resolution[1],
    y = 1:resolution[2]
  ) %>%
    mutate(effort = effort_vec)

  ggplot(patch_effort) +
    geom_tile(aes(x, y, fill = effort)) +
    geom_point(data = ports, aes(x, y),
               color = "red", size = 3, shape = 17) +
    scale_fill_viridis_c(option = "C") +
    labs(title = title, fill = "Effort") +
    coord_equal() +
    theme(plot.title = element_text(size = 11))
}

plot_biomass <- function(proc, title = "") {
  last_step <- max(proc$fauna$step)

  ssb_df <- proc$fauna %>%
    filter(step == last_step) %>%
    group_by(critter, x, y) %>%
    summarise(ssb = sum(ssb, na.rm = TRUE), .groups = "drop")

  ggplot(ssb_df) +
    geom_tile(aes(x, y, fill = ssb)) +
    geom_point(data = ports, aes(x, y),
               color = "red", size = 2, shape = 17) +
    scale_fill_viridis_c(option = "B") +
    facet_wrap(~critter) +
    coord_equal() +
    labs(title = title, fill = "SSB") +
    theme(plot.title = element_text(size = 11))
}
```

## Scenario 1: `rpue` vs `revenue`

In this scenario the longline fleet allocates effort based on `rpue` (revenue per unit effort) while the handline fleet chases total `revenue`.

The `rpue` fleet moves toward patches where catch rates are highest per unit effort. This tends to spread effort more evenly, because as a high-RPUE patch attracts more effort, catch rates decline there relative to less-fished patches.

The `revenue` fleet, by contrast, gravitates toward patches with the highest *total* revenue regardless of how much effort is already there. This can lead to stronger concentration of effort in productive patches, since high effort × moderate catch rate can still produce high total revenue.

Neither strategy accounts for travel costs, so port location does not directly influence where these fleets fish.

```{r scenario-1}
fleets_1 <- build_fleets("rpue", "revenue")

time_1 <- system.time({
  sim_1 <- simmar(fauna = fauna, fleets = fleets_1, years = years)
})

proc_1 <- process_marlin(sim_1, time_step = time_step)

plot_effort(sim_1, "longline", "Longline Effort (rpue)")
plot_effort(sim_1, "handline", "Handline Effort (revenue)")
plot_biomass(proc_1, "SSB — rpue vs revenue")
```

Notice how the longline fleet (rpue) distributes effort more diffusely across productive patches, while the handline fleet (revenue) concentrates more heavily in the highest-biomass areas. The biomass maps show the resulting depletion patterns: species are drawn down most where their habitat overlaps with each fleet's effort hotspots.


## Scenario 2: `ppue` vs `profit`

Now we contrast the two cost-aware strategies. The longline fleet uses `ppue` (profit per unit effort) and the handline fleet uses total `profit`. Both account for the cost of traveling from port, but they differ in whether the objective is per-unit-effort or total.

Because `ppue` divides profit by effort, it penalizes patches where effort is high relative to returns. The `profit` strategy rewards patches where total profit (revenue minus costs) is maximized, which can favor concentrating effort in nearby productive patches even if per-unit returns are declining.

With `travel_fraction = 0.3` and two asymmetrically placed ports, we should see both fleets pulled toward the ports relative to the cost-naive strategies above.

```{r scenario-2}
fleets_2 <- build_fleets("ppue", "profit")

time_2 <- system.time({
  sim_2 <- simmar(fauna = fauna, fleets = fleets_2, years = years)
})

proc_2 <- process_marlin(sim_2, time_step = time_step)

plot_effort(sim_2, "longline", "Longline Effort (ppue)")
plot_effort(sim_2, "handline", "Handline Effort (profit)")
plot_biomass(proc_2, "SSB — ppue vs profit")
```

Compare these maps to Scenario 1. The cost-aware fleets are pulled toward the port locations (red triangles), and patches far from any port receive less effort than they would under `rpue` or `revenue`. The biomass maps show that remote patches retain more biomass under cost-aware allocation, since fleets fish them less intensively.


## Scenario 3: `marginal_profit` vs `rpue`

Finally, we demonstrate `marginal_profit`, the most economically sophisticated allocation strategy. The longline fleet now allocates based on the *marginal return* to an additional unit of effort in each patch, while the handline fleet uses the simpler `rpue` as a baseline for comparison.

Under `marginal_profit`, the fleet responds to diminishing returns: even if a patch has high total profit or high RPUE, if the *next* unit of effort there would yield little additional return, the fleet reallocates toward patches where marginal returns are still high. This tends to produce the most even spatial distribution of effort among all the profit-aware strategies, because it directly penalizes over-concentration.

```{r scenario-3}
fleets_3 <- build_fleets("marginal_profit", "rpue")

time_3 <- system.time({
  sim_3 <- simmar(fauna = fauna, fleets = fleets_3, years = years)
})

proc_3 <- process_marlin(sim_3, time_step = time_step)

plot_effort(sim_3, "longline", "Longline Effort (marginal_profit)")
plot_effort(sim_3, "handline", "Handline Effort (rpue)")
plot_biomass(proc_3, "SSB — marginal_profit vs rpue")
```

The `marginal_profit` fleet (longline) typically produces a smoother effort surface than any of the other cost-aware strategies. Compared to the `rpue` handline fleet sharing the same waters, the marginal profit fleet is more sensitive to congestion: it avoids piling into patches that are already heavily fished, even if those patches still have high average returns. The biomass maps reflect this — depletion tends to be more spatially uniform under `marginal_profit`.

## Scenario 4: `sole_owner` vs `open_access`

The previous scenarios varied the *spatial* allocation strategy while keeping the fleet model (how total effort changes over time) fixed. Now we flip that: both fleets use the same spatial allocation (`ppue`) but differ in their *fleet model*.

The classic result from fisheries economics is that open access fleets fish until average profit is zero (all rents dissipated), while a sole owner — who internalizes the stock externality — stops expanding effort when the *marginal* profit of the next unit of effort hits zero. This corresponds to maximum economic yield (MEY), which occurs at lower effort and higher biomass than the open access equilibrium.

In `marlin`, `fleet_model = "sole_owner"` implements this by using the same dynamic entry/exit machinery as open access, but replacing the profitability signal: instead of responding to total profits, the sole owner responds to the effort-weighted mean of patch-level marginal profits computed by `calc_marginal_value`.

We build this scenario manually rather than using `build_fleets`, since the fleets differ in `fleet_model` rather than `spatial_allocation`.

```{r scenario-4}
fleets_oa <- list(
  "longline" = create_fleet(
    list(
      "bigeye" = Metier$new(
        critter = fauna$bigeye,
        price = 10,
        sel_form = "logistic",
        sel_start = 1,
        sel_delta = 0.01,
        catchability = 0,
        p_explt = 2
      ),
      "skipjack" = Metier$new(
        critter = fauna$skipjack,
        price = 5,
        sel_form = "logistic",
        sel_start = 0.8,
        sel_delta = 0.1,
        catchability = 0,
        p_explt = 1
      )
    ),
    ports = ports,
    base_effort = patches,
    resolution = resolution,
    spatial_allocation = "ppue",
    fleet_model = "open_access",
    cr_ratio = 1,
    travel_fraction = 0.7
  ),
  "handline" = create_fleet(
    list(
      "bigeye" = Metier$new(
        critter = fauna$bigeye,
        price = 10,
        sel_form = "logistic",
        sel_start = 1.2,
        sel_delta = 0.1,
        catchability = 0,
        p_explt = 1
      ),
      "skipjack" = Metier$new(
        critter = fauna$skipjack,
        price = 8,
        sel_form = "logistic",
        sel_start = 0.5,
        sel_delta = 0.2,
        catchability = 0,
        p_explt = 2
      )
    ),
    ports = ports,
    base_effort = patches,
    resolution = resolution,
    spatial_allocation = "ppue",
    fleet_model = "open_access",
    cr_ratio = 1,
    travel_fraction = 0.5
  )
)

fleets_so <- fleets_oa
fleets_so$longline$fleet_model <- "sole_owner"
fleets_so$handline$fleet_model <- "sole_owner"

fleets_oa <- tune_fleets(fauna, fleets_oa, tune_type = "depletion")
fleets_so <- tune_fleets(fauna, fleets_so, tune_type = "depletion")
```

Now we run both simulations side by side.

```{r scenario-4-run}
years_4 <- 50  # longer run to let both fleet models reach equilibrium

time_4a <- system.time({
  sim_4a <- simmar(fauna = fauna, fleets = fleets_oa, years = years_4)
})

time_4b <- system.time({
  sim_4b <- simmar(fauna = fauna, fleets = fleets_so, years = years_4)
})

proc_4a <- process_marlin(sim_4a, time_step = time_step)
proc_4b <- process_marlin(sim_4b, time_step = time_step)
```

The key comparison is the trajectory of total effort and profits over time. The open access fleet should converge to zero profits, while the sole owner should stabilize at lower effort with positive profits.

```{r scenario-4-effort-time, fig.width = 9, fig.height = 5}
effort_ts <- bind_rows(
  proc_4a$fleets %>%
    group_by(step, fleet) %>%
    summarise(effort = sum(effort), .groups = "drop") %>%
    mutate(model = "open_access"),
  proc_4b$fleets %>%
    group_by(step, fleet) %>%
    summarise(effort = sum(effort), .groups = "drop") %>%
    mutate(model = "sole_owner")
)

ggplot(effort_ts, aes(step * time_step, effort, color = model)) +
  geom_line() +
  facet_wrap(~fleet) +
  scale_x_continuous(name = "Year") +
  scale_y_continuous(name = "Total Effort", limits = c(0, NA)) +
  labs(title = "Total Effort: Open Access vs Sole Owner",
       color = "Fleet Model")
```

The sole owner settles at lower effort than open access for both fleets, because it stops expanding when marginal returns hit zero rather than waiting for average returns to be exhausted.

We can also compare profits over time. Open access should converge toward zero, while the sole owner retains positive profits at equilibrium (the hallmark of MEY).

```{r scenario-4-profits, fig.width = 9, fig.height = 5}
profit_ts <- bind_rows(
  proc_4a$fleets %>%
    group_by(step, fleet) %>%
    summarise(revenue = sum(revenue), catch = sum(catch), .groups = "drop") %>%
    mutate(model = "open_access"),
  proc_4b$fleets %>%
    group_by(step, fleet) %>%
    summarise(revenue = sum(revenue), catch = sum(catch), .groups = "drop") %>%
    mutate(model = "sole_owner")
)

ggplot(profit_ts, aes(step * time_step, revenue, color = model)) +
  geom_line() +
  facet_wrap(~fleet) +
  scale_x_continuous(name = "Year") +
  labs(title = "Revenue: Open Access vs Sole Owner",
       color = "Fleet Model",
       y = "Revenue")
```

Finally, compare the spatial effort and biomass footprints at the end of each simulation.

```{r scenario-4-space}
plot_effort(sim_4a, "longline", "Longline Effort (open_access)")
plot_effort(sim_4b, "longline", "Longline Effort (sole_owner)")
plot_biomass(proc_4a, "SSB — open_access")
plot_biomass(proc_4b, "SSB — sole_owner")
```

The sole owner leaves more biomass in the water because it fishes less intensively. This is the fundamental tradeoff: open access maximizes effort and dissipates rents, while the sole owner restrains effort to maximize total profit, resulting in higher biomass and positive economic returns.

## Scenario 5: spatial-allocation memory

The previous scenarios all show smooth, well-behaved dynamics — even after the MPA closure in Scenario 4. The one-step-lag feedback that motivates `objective_memory_halflife` only bites strongly under specific configurations. The most reliable way to surface it is with a `marginal_profit` allocator: the marginal value of effort at a patch responds directly to how much was fished there last step, so a single-step lag can produce period-2 sawtooth in patch effort and (faintly) in catch.

Here we set up a 2-fleet system with both fleets using `marginal_profit` allocation and close an offshore corner mid-run, then sweep `objective_memory_halflife` across a small range. The expectation: at `halflife = 0` (legacy behavior) the dynamics show visible high-frequency wiggle in individual patches; modest memory dampens it cleanly; very long memory introduces a different problem — low-frequency overshoot driven by the phase lag the smoothing imposes.

```{r scenario-5-setup}
years_5    <- 20
mpa_year_5 <- 8

# Close an offshore corner starting at year 8
mpa_locs_5 <- expand_grid(x = 1:resolution[1], y = 1:resolution[2]) %>%
  mutate(mpa = x >= 12 & y >= 12)
manager_5 <- list(mpas = list(locations = mpa_locs_5, mpa_year = mpa_year_5))

# Fleet builder parameterised on a single shared halflife (both fleets get it).
# Uses the same metier / fleet structure as Scenarios 1-3.
build_memory_fleets <- function(halflife) {
  fleets <- list(
    "longline" = create_fleet(
      list(
        "bigeye" = Metier$new(
          critter = fauna$bigeye, price = 10,
          sel_form = "logistic", sel_start = 1, sel_delta = 0.01,
          catchability = 0, p_explt = 2
        ),
        "skipjack" = Metier$new(
          critter = fauna$skipjack, price = 5,
          sel_form = "logistic", sel_start = 0.8, sel_delta = 0.1,
          catchability = 0, p_explt = 1
        )
      ),
      ports = ports, base_effort = patches, resolution = resolution,
      spatial_allocation = "marginal_profit",
      fleet_model = "constant_effort",
      cr_ratio = 1, travel_fraction = 0.7,
      objective_memory_halflife = halflife
    ),
    "handline" = create_fleet(
      list(
        "bigeye" = Metier$new(
          critter = fauna$bigeye, price = 10,
          sel_form = "logistic", sel_start = 1.2, sel_delta = 0.1,
          catchability = 0, p_explt = 1
        ),
        "skipjack" = Metier$new(
          critter = fauna$skipjack, price = 8,
          sel_form = "logistic", sel_start = 0.5, sel_delta = 0.2,
          catchability = 0, p_explt = 2
        )
      ),
      ports = ports, base_effort = patches, resolution = resolution,
      spatial_allocation = "marginal_profit",
      fleet_model = "constant_effort",
      cr_ratio = 1, travel_fraction = 0.5,
      objective_memory_halflife = halflife
    )
  )
  tune_fleets(fauna, fleets, tune_type = "depletion")
}
```

```{r scenario-5-run}
halflives_5 <- c(0, 1, 2)

runs_5 <- lapply(halflives_5, function(hl) {
  sim  <- simmar(fauna = fauna, fleets = build_memory_fleets(hl),
                 years = years_5, manager = manager_5)
  proc <- process_marlin(sim, time_step = time_step)
  list(halflife = hl, proc = proc)
})
```

Two patch-level metrics, computed on the post-MPA window:

- **lag-1 autocorrelation of patch effort** (low / negative ⇒ sawtooth)
- **step-difference CV**: `sd(diff(effort)) / mean(effort)`, capturing high-frequency wiggle on top of any trend

```{r scenario-5-metrics}
post_mpa_metrics <- function(r, burn_after_mpa_years = 2) {
  start_step <- (mpa_year_5 + burn_after_mpa_years) * seasons
  r$proc$fleets %>%
    filter(step > start_step) %>%
    arrange(fleet, patch, step) %>%
    group_by(fleet, patch) %>%
    summarise(
      mean_effort  = mean(effort, na.rm = TRUE),
      effort_acf1  = if (n() > 3 && stats::sd(effort, na.rm = TRUE) > 0)
        stats::acf(effort, plot = FALSE, lag.max = 1)$acf[2] else NA_real_,
      step_diff_cv = if (n() > 3 && mean(effort, na.rm = TRUE) > 1e-9)
        stats::sd(diff(effort), na.rm = TRUE) / mean(effort, na.rm = TRUE) else NA_real_,
      .groups = "drop"
    ) %>%
    mutate(halflife = r$halflife)
}

all_metrics <- purrr::map_dfr(runs_5, post_mpa_metrics) %>%
  filter(mean_effort > 1e-3)

all_metrics %>%
  group_by(halflife, fleet) %>%
  summarise(
    n_patches            = n(),
    median_effort_acf1   = signif(median(effort_acf1, na.rm = TRUE), 3),
    median_step_diff_cv  = signif(median(step_diff_cv, na.rm = TRUE), 3),
    p90_step_diff_cv     = signif(quantile(step_diff_cv, 0.9, na.rm = TRUE), 3),
    .groups = "drop"
  ) %>%
  knitr::kable(digits = 3,
               caption = "Post-MPA per-patch effort autocorrelation and high-frequency wiggle, by halflife and fleet.")
```

```{r scenario-5-traces, fig.height = 6}
# Pick the 3 most-oscillating patches per fleet under the lowest halflife;
# trace each across all three runs.
baseline_hl  <- min(halflives_5)
top_patches  <- all_metrics %>%
  filter(halflife == baseline_hl) %>%
  group_by(fleet) %>%
  slice_max(step_diff_cv, n = 3, with_ties = FALSE) %>%
  ungroup() %>%
  select(fleet, patch)

trace_df <- purrr::map_dfr(runs_5, function(r) {
  r$proc$fleets %>%
    semi_join(top_patches, by = c("fleet", "patch")) %>%
    transmute(fleet, patch, step, year = step * time_step, effort,
              halflife = factor(r$halflife, levels = sort(unique(halflives_5))))
})

ggplot(trace_df, aes(year, effort, color = halflife)) +
  geom_line() +
  geom_vline(xintercept = mpa_year_5, linetype = "dashed", alpha = 0.5) +
  facet_grid(fleet ~ patch, scales = "free_y", labeller = label_both) +
  scale_color_viridis_d(option = "C", end = 0.85) +
  labs(
    title    = "Patch effort traces, most-oscillating patches per fleet",
    subtitle = sprintf("Dashed line: MPA closure at year %d. halflife is years (season-agnostic).",
                       mpa_year_5),
    x = "Year", y = "Patch effort", color = "halflife (yr)"
  )
```

```{r scenario-5-catch, fig.height = 4}
catch_ts <- purrr::map_dfr(runs_5, function(r) {
  r$proc$fauna %>%
    group_by(step) %>%
    summarise(catch = sum(c, na.rm = TRUE), .groups = "drop") %>%
    mutate(year = step * time_step,
           halflife = factor(r$halflife, levels = sort(unique(halflives_5))))
})

ggplot(catch_ts, aes(year, catch, color = halflife)) +
  geom_line() +
  geom_vline(xintercept = mpa_year_5, linetype = "dashed", alpha = 0.5) +
  scale_color_viridis_d(option = "C", end = 0.85) +
  labs(title = "Total catch over time",
       x = "Year", y = "Total catch", color = "halflife (yr)")
```

Reading the patch traces: `halflife = 0` shows the high-frequency wiggle that motivates the parameter, sharpened further by the MPA closure. `halflife = 1` (one year of memory) sits near the median of the sawtooth and removes the high-frequency component. `halflife = 2` smooths further but begins to introduce slower swings driven by phase lag: the fleet's perceived objective is now lagging the real one by ~3 years, so it reallocates toward patches that *used to be* high-value.

### When does memory matter?

The susceptibility to one-step-lag oscillation depends strongly on a fleet's configuration. Memory is not a universal upgrade; in many configurations it adds nothing or hurts.

- **Allocation strategy.** `marginal_profit` and `marginal_revenue` have the most direct fleet-biomass feedback — the objective at a patch responds immediately to last step's fishing there — and are the most prone to sawtooth. `ppue` / `profit` / `rpue` / `revenue` integrate over the production curve, which damps the feedback. `manual` and `uniform` ignore the buffet entirely, so memory does nothing for them (and is correctly skipped by `simmar()`).
- **Fleet model.** `constant_effort` fleets have only the spatial-allocation feedback loop, so the spatial smoothing in memory is the only mechanism stabilising them. `open_access` and `sole_owner` fleets also adjust *total* effort each step from a fleet-wide profitability signal, which has its own slow damping built in via the annual entry/exit caps — but that signal itself is **not** smoothed by `objective_memory_halflife`. If you observe sawtooth in total fleet effort (rather than in patch-level allocation), memory will not fix it; the right knob lives in the open-access / sole-owner parameters.
- **System structure.** Strong fleet-biomass coupling (high catchability relative to stock productivity, fast movement that smooths biomass and equalises marginals, MPA closures that force redistribution) all amplify the one-step-lag feedback. Without any of these, the canonical scenarios above behave smoothly even at `halflife = 0`.

Sensible values for `objective_memory_halflife` depend on the time scale of stock and fleet dynamics. For most marine fisheries with annual or sub-annual seasons, **0.5–1.5 years** is a useful starting range. If catch or effort trajectories at a chosen halflife show slow swings that aren't present at `halflife = 0`, reduce it; you've crossed into the phase-lag regime. If the system is well-behaved at `halflife = 0`, leave it there.

## Runtime Comparison

The `marginal_profit` spatial allocation strategy and the `sole_owner` fleet model both require computing numerical derivatives of the profit function via `calc_marginal_value` at every time step, which adds extra `go_fish` evaluations (one per fleet at minimum). This makes them meaningfully slower than strategies that only need the quantities already computed by the standard `go_fish` call.

```{r runtime}
runtime <- tibble(
  scenario = c("rpue / revenue", "ppue / profit", "marginal_profit / rpue",
               "open_access (ppue)", "sole_owner (ppue)"),
  seconds = c(time_1["elapsed"], time_2["elapsed"], time_3["elapsed"],
              time_4a["elapsed"], time_4b["elapsed"]),
  years_run = c(rep(years, 3), years_4, years_4)
) %>%
  mutate(
    sec_per_year = round(seconds / years_run, 2),
    relative = round(sec_per_year / min(sec_per_year), 1)
  )

knitr::kable(runtime,
             col.names = c("Scenario", "Seconds", "Years", "Sec/Year", "Relative to fastest"),
             digits = 2,
             caption = "Wall-clock time for simmar() by scenario. Sec/Year normalizes for different run lengths.")
```

## Summary Comparison

To see all five spatial allocation strategies side by side, we extract final-step effort across Scenarios 1–3 and compare in a single figure.

```{r comparison-effort, fig.width = 10, fig.height = 10}
extract_effort <- function(sim, fleet_name, label) {
  final_step <- sim[[length(sim)]]
  effort_vec <- final_step[[1]]$e_p_fl[, fleet_name]

  expand_grid(x = 1:resolution[1], y = 1:resolution[2]) %>%
    mutate(
      effort = effort_vec,
      strategy = label
    )
}

all_efforts <- bind_rows(
  extract_effort(sim_1, "longline", "rpue"),
  extract_effort(sim_1, "handline", "revenue"),
  extract_effort(sim_2, "longline", "ppue"),
  extract_effort(sim_2, "handline", "profit"),
  extract_effort(sim_3, "longline", "marginal_profit")
) %>%
  mutate(strategy = factor(strategy,
    levels = c("rpue", "revenue", "ppue", "profit", "marginal_profit")))

ggplot(all_efforts) +
  geom_tile(aes(x, y, fill = effort)) +
  geom_point(data = ports, aes(x, y),
             color = "red", size = 2, shape = 17) +
  scale_fill_viridis_c(option = "C") +
  facet_wrap(~strategy, ncol = 3) +
  coord_equal() +
  labs(
    title = "Effort Distribution by Spatial Allocation Strategy",
    subtitle = "Red triangles = port locations",
    fill = "Effort"
  )
```

We can do the same for SSB to see how each allocation strategy shapes the resulting biomass landscape. The rows correspond to the fleet allocation pairings (longline / handline), and the columns to each species.

```{r comparison-ssb, fig.width = 10, fig.height = 12}
extract_ssb <- function(proc, longline_alloc, handline_alloc) {
  last_step <- max(proc$fauna$step)

  proc$fauna %>%
    filter(step == last_step) %>%
    group_by(critter, x, y) %>%
    summarise(ssb = sum(ssb, na.rm = TRUE), .groups = "drop") %>%
    mutate(strategy = paste0(longline_alloc, " / ", handline_alloc))
}

all_ssb <- bind_rows(
  extract_ssb(proc_1, "rpue", "revenue"),
  extract_ssb(proc_2, "ppue", "profit"),
  extract_ssb(proc_3, "marginal_profit", "rpue")
)

ggplot(all_ssb) +
  geom_tile(aes(x, y, fill = ssb)) +
  geom_point(data = ports, aes(x, y),
             color = "red", size = 1.5, shape = 17) +
  scale_fill_viridis_c(option = "B") +
  facet_grid(strategy ~ critter) +
  coord_equal() +
  labs(
    title = "SSB by Allocation Strategy (longline / handline)",
    subtitle = "Red triangles = port locations",
    fill = "SSB"
  )
```


Some patterns to look for across these panels:

- **`rpue` vs `revenue`**: The per-unit strategy (`rpue`) tends to spread effort more evenly, while the total strategy (`revenue`) concentrates effort in high-biomass patches. Since neither accounts for travel costs, effort extends freely into remote patches.
- **`ppue` vs `profit`**: Adding travel costs shifts effort toward ports. The per-unit version (`ppue`) is especially reluctant to send effort far from port, creating biomass refugia in remote areas.
- **`marginal_profit`**: Produces the smoothest effort surface among cost-aware strategies, because it directly penalizes diminishing returns from effort concentration. This leads to the most spatially uniform depletion.
- **`sole_owner` vs `open_access`**: With the same spatial allocation (`ppue`), the sole owner reaches equilibrium at lower total effort and higher biomass than open access, retaining positive profits where open access dissipates them. This is the classic MEY vs. open-access result, now playing out in a spatially explicit setting.

The choice of `spatial_allocation` can substantially change the spatial footprint of fishing, which in turn affects local depletion patterns, the effectiveness of spatial management tools like MPAs, and the economic outcomes of the fishery. See the `port-distance` and `fleet-management` vignettes for more on how these fleet options interact with other `marlin` features.
